---
title: "Recall: ggplot2 and dplyr"
format:
  revealjs:
    width: 1280
    height: 720
    margin: 0.02
    smaller: true
    echo: true
    message: false
    fig-width: 7
    fig-height: 7
---

```{css}
div.sourceCode {
  font-size: 1.4em;
}

div.cell-output-stdout {
  font-size: 1.4em;
}
```

##

```{r}
#| setup: true

library(tidyverse)
theme_set(theme_gray(base_size=16))
```

![](figures/rnz_sheep_vs_humans)

[https://www.rnz.co.nz/news/country/560252/gap-between-people-and-sheep-rapidly-closing](https://www.rnz.co.nz/news/country/560252/gap-between-people-and-sheep-rapidly-closing)

## Data for today

[Figure.nz](https://figure.nz) has data on agriculture production in Aotearoa NZ from 1935 and population from 1937:

[https://figure.nz/table/NteyKZu48ZHuqab1](https://figure.nz/table/NteyKZu48ZHuqab1)
[https://figure.nz/table/83PJnWFeJehqJYaS](https://figure.nz/table/83PJnWFeJehqJYaS)

::: aside
These URLs may not be stable over time.
:::

## Data for today

```{r}
library(tidyverse)

ag_production <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/nz/ag_production.csv")
ag_production
```

## Data for today

```{r}
#| message: false

population <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/nz/nz_population.csv")
population
```

## The ggplot2 recipe

1. Create a chart with `ggplot2()`
2. Assign some data to the chart with `data=foo`, i.e. `ggplot2(data=foo)`
3. Define the mapping from features of the chart to variables/columns in the data with `aes()`
4. Choose which type of geometry to use with `geom_*()`.

## The ggplot2 recipe

:::: {.columns}

::: {.column width="50%"}
```{r gg1}
#| eval: false
ggplot()
```

`ggplot()` creates a new chart
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "gg1"
```
:::

::::


## The ggplot2 recipe

:::: {.columns}

::: {.column width="50%"}
```{r gg2}
#| eval: false
ggplot(data=ag_production)
```

`ggplot()` creates a new chart

`data=ag_production` assigns the data to the chart
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "gg2"
```
:::

::::

## The ggplot2 recipe

:::: {.columns}

::: {.column width="50%"}
```{r gg3}
#| eval: false
ggplot(data=ag_production) +
  aes(x=year, y=value)
```

`ggplot()` creates a new chart

`data=ag_production` assigns the data to the chart

`aes(x=year, y=value)` maps features of chart to columns of data
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "gg3"
```
:::

::::

## The ggplot2 recipe

:::: {.columns}

::: {.column width="50%"}
```{r gg4}
#| eval: false
ggplot(data=ag_production) +
  aes(x=year, y=value) +
  geom_point()
```

`ggplot()` creates a new chart

`data=ag_production` assigns the data to the chart

`aes(x=year, y=value)` maps features of chart to columns of data

`geom_point()` chooses the point geometry
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "gg4"
```
:::

::::

## Adding another aesthetic

:::: {.columns}

::: {.column width="50%"}
```{r gg5}
#| eval: false
ggplot(data=ag_production) +
  aes(x=year, y=value, colour=animal) +
  geom_point()
```

`colour=animal` maps the colour aesthetic to the `animal` column.

Note it automatically supplies a *scale*, in this case a legend or **guide**.

`x` and `y` also have scales!
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "gg5"
```
:::

::::

## A different geometry

:::: {.columns}

::: {.column width="50%"}
```{r gg6}
#| eval: false
ggplot(data=ag_production) +
  aes(x=year, y=value, colour=animal) +
  geom_line()
```

The line geometry draws lines between observations within 'groups'.

The grouping is automatic when you have assigned `colour` or `fill`.
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "gg6"
```
:::

::::

## Groups are important

:::: {.columns}

::: {.column width="50%"}
```{r gg7}
#| eval: false
ggplot(data=ag_production) +
  aes(x=year, y=value) +
  geom_line()
```

If there are no groups, all lines will be joined up.
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "gg7"
```
:::

::::


::::

## Groups are important

:::: {.columns}

::: {.column width="50%"}
```{r gg8}
#| eval: false
ggplot(data=ag_production) +
  aes(x=year, y=value, group=animal) +
  geom_line()
```

The `group` aesthetic can be useful in this case.

But it has no scale or guide.
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "gg8"
```
:::

::::

## Labelling

:::: {.columns}

::: {.column width="50%"}
```{r gg9}
#| eval: false
ggplot(data=ag_production) +
  aes(x=year, y=value, colour=animal) +
  geom_line() +
  labs(x = "Year",
       y = "Number of animals",
       colour = "Animal")
```

The `labs()` command allows you to override the default labels.

The parameters it takes are the same as aesthetics.

:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "gg9"
```
:::

::::

## Label dictionary

:::: {.columns}

::: {.column width="50%"}
```{r gg10}
#| eval: false

labels = c(year = "Year",
           animal = "Animal",
           value = "Number of Animals")

ggplot(data=ag_production) +
  aes(x=year, y=value, colour=animal) +
  geom_line() +
  labs(dictionary = labels)
```

A label dictionary is a named vector mapping the columns
in your data to the label you want to use.

This can be useful for consistency across multiple plots.
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "gg10"
```
:::

::::

::: aside
Recall the `c()` function combines multiple entries into a single vector.
:::

## The grammar of graphics

ggplot2 is based on the Grammar of Graphics defined by Leland Wilkinson in 1999.

In the same way that a grammar defines the structures and composition of a language, the grammar of graphics defines the structures and composition of statistical graphics.

Most widely used graphics packages for charting have been informed by this idea.

The grammar in ggplot2 breaks down a chart into:

- Data
- Aesthetics
- Geometries
- Facets
- Statistics
- Coordinates
- Themes

## Data

The data 'layer' in `ggplot2` needs to be **tidy**. That is, you need to ensure that each observation is in its own row and each variable is in its own column.

The `ag_production` data are tidy in **long form**: each row represents a single observation the count (`value`) of `animal` in a particular `year`.

However, we could also present these data in **wide form** where each animal has its own column instead:

```{r}
ag_wide <- ag_production |> pivot_wider(names_from=animal, values_from=value) |>
  filter(year > 1970)
```

It's the same data, just formatted the other way around. For charting though, it probably doesn't make sense to plot the `Sheep` column against the `Goats` column directly: both are changing through time, and having time be on the `x` axis is likely sensible!

## Aesthetics

Aesthetics are features of the plot that allow us to distinguish one observation from another.

We've already seen position aesthetics (`x` and `y`) as well as the `colour` and `group` aesthetics. Others are `shape`, `size`, `alpha` (transparency) `linewidth` and `linetype`.

## Aesthetics

:::: {.columns}

::: {.column width="50%"}
```{r gg11}
#| eval: false
ggplot(data=ag_production) +
  aes(x=year, y=value, shape=animal) +
  geom_point()
```

`shape` can be useful when we have a relatively small number of groups.
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "gg11"
```
:::

::::


## Aesthetics

:::: {.columns}

::: {.column width="50%"}
```{r gg12}
#| eval: false
ggplot(data=ag_production) +
  aes(x=year, y=value, size=animal) +
  geom_point()
```

`size` is typically a continuous measure rather than a discrete one.
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "gg12"
```
:::

::::


## Aesthetics

:::: {.columns}

::: {.column width="50%"}
```{r gg13}
#| eval: false
ggplot(data=ag_production) +
  aes(x=year, y=value, colour=animal, shape=animal) +
  geom_point()
```

You can combine aesthetics and the guides will try and combine as well.
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "gg13"
```
:::

::::


## Aesthetics

:::: {.columns}

::: {.column width="50%"}
```{r gg14}
#| eval: false
ggplot(data=ag_production) +
  aes(x=year, y=value, linewidth=animal) +
  geom_line()
```

Line geometries can be distinguished using width and line type.
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "gg14"
```
:::

::::


## Aesthetics

:::: {.columns}

::: {.column width="50%"}
```{r gg15}
#| eval: false
ggplot(data=ag_production) +
  aes(x=year, y=value, linetype=animal) +
  geom_line()
```

Line geometries can be distinguished using width and line type.
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "gg15"
```
:::

::::
