---
title: "Recall: ggplot2 and dplyr"
format:
  revealjs:
    width: 1280
    height: 720
    margin: 0.02
    smaller: true
    echo: true
    message: false
    fig-width: 7
    fig-height: 7
---

```{css}
/*| echo: false */

div.sourceCode {
  font-size: 1.4em;
}

div.cell-output-stdout {
  font-size: 1.4em;
}
```

```{r}
#| echo: false

library(tidyverse)
theme_set(theme_gray(base_size=16))
```

![](figures/rnz_sheep_vs_humans)

[https://www.rnz.co.nz/news/country/560252/gap-between-people-and-sheep-rapidly-closing](https://www.rnz.co.nz/news/country/560252/gap-between-people-and-sheep-rapidly-closing)

## Data for today

[Figure.nz](https://figure.nz) has data on agriculture production in Aotearoa NZ from 1935:

[https://figure.nz/table/NteyKZu48ZHuqab1](https://figure.nz/table/NteyKZu48ZHuqab1)

And also population data from 1937:

[https://figure.nz/table/83PJnWFeJehqJYaS](https://figure.nz/table/83PJnWFeJehqJYaS)

::: aside
These URLs may not be stable over time.
:::

## Data for today

```{r}
library(tidyverse)

ag_production <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/nz/ag_production.csv")
ag_production
```

## Data for today

```{r}
#| message: false

population <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/nz/nz_population.csv")
population
```

##

```{r sheep_vs_humans_final}
#| echo: false
#| fig.width: 16
#| fig.height: 9
ag_production |>
  left_join(population, by=join_by(year)) |>
  filter(animal == "Sheep",
         !is.na(population)) |>
  mutate(ratio = value/population) |>
  ggplot() +
  aes(x=year, y=ratio) +
  geom_line(col = "purple", linewidth=2) +
  geom_hline(yintercept = 1, linewidth=1, linetype='dotted', col='grey40') +
  annotate(geom='text', x=1935, y=1, label='1 sheep per person', vjust=-0.3, col='grey40', hjust=0) +
  scale_y_continuous(limits = \(y) range(y, 1)) +
  labs(x = NULL,
       y = "Sheep per person",
       title = "Gap between people and sheep rapidly closing") +
  theme_minimal(base_size=32)
```

# ggplot2

## The ggplot2 recipe

1. Create a chart with `ggplot2()`
2. Assign some data to the chart with `data=foo`, i.e. `ggplot2(data=foo)`
3. Define the mapping from features of the chart to variables/columns in the data with `aes()`
4. Choose which type of geometry to use with `geom_*()`.

## The ggplot2 recipe

:::: {.columns}

::: {.column width="50%"}
```{r gg1}
#| eval: false
ggplot()
```

`ggplot()` creates a new chart
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "gg1"
```
:::

::::


## The ggplot2 recipe

:::: {.columns}

::: {.column width="50%"}
```{r gg2}
#| eval: false
ggplot(data=ag_production)
```

`ggplot()` creates a new chart

`data=ag_production` assigns the data to the chart
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "gg2"
```
:::

::::

## The ggplot2 recipe

:::: {.columns}

::: {.column width="50%"}
```{r gg3}
#| eval: false
ggplot(data=ag_production) +
  aes(x=year, y=value)
```

`ggplot()` creates a new chart

`data=ag_production` assigns the data to the chart

`aes(x=year, y=value)` maps features of chart to columns of data
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "gg3"
```
:::

::::

## The ggplot2 recipe

:::: {.columns}

::: {.column width="50%"}
```{r gg4}
#| eval: false
ggplot(data=ag_production) +
  aes(x=year, y=value) +
  geom_point()
```

`ggplot()` creates a new chart

`data=ag_production` assigns the data to the chart

`aes(x=year, y=value)` maps features of chart to columns of data

`geom_point()` chooses the point geometry
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "gg4"
```
:::

::::

## Adding another aesthetic

:::: {.columns}

::: {.column width="50%"}
```{r gg5}
#| eval: false
ggplot(data=ag_production) +
  aes(x=year, y=value, colour=animal) +
  geom_point()
```

`colour=animal` maps the colour aesthetic to the `animal` column.

Note it automatically supplies a **scale**, in this case a legend or **guide**.

`x` and `y` also have scales!
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "gg5"
```
:::

::::

## A different geometry

:::: {.columns}

::: {.column width="50%"}
```{r gg6}
#| eval: false
ggplot(data=ag_production) +
  aes(x=year, y=value, colour=animal) +
  geom_line()
```

The line geometry draws lines between observations within 'groups'.

The grouping is automatic when you have assigned `colour` or `fill`.
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "gg6"
```
:::

::::

## Groups are important

:::: {.columns}

::: {.column width="50%"}
```{r gg7}
#| eval: false
ggplot(data=ag_production) +
  aes(x=year, y=value) +
  geom_line()
```

If there are no groups, all lines will be joined up.
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "gg7"
```
:::

::::

## Groups are important

:::: {.columns}

::: {.column width="50%"}
```{r gg8}
#| eval: false
ggplot(data=ag_production) +
  aes(x=year, y=value, group=animal) +
  geom_line()
```

The `group` aesthetic can be useful in this case.

But it has no scale or guide.
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "gg8"
```
:::

::::

## The grammar of graphics

ggplot2 is based on the Grammar of Graphics defined by Leland Wilkinson in 1999.

In the same way that a grammar defines the structures and composition of a language, the grammar of graphics defines the structures and composition of statistical graphics.

Most widely used graphics packages for charting have been informed by this idea.

::: {.fragment}
The grammar in ggplot2 breaks down a chart into:

- Data
- Aesthetics
- Geometries
- Facets
- Statistics
- Coordinates
- Themes
:::

## Data

The data 'layer' in `ggplot2` needs to be **tidy**. That is, you need to ensure that

- each observation is in its own row and
- each variable is in its own column.

::: {.fragment}
The `ag_production` data are tidy:

- there are three variables (`year`, `animal`, `value`) in separate columns.
- each row represents a single observation: the count (`value`) of `animal` in a particular `year`.
:::

## Data

We could also present these data in **wide form** where each animal has its own column instead:

```{r}
ag_production |> pivot_wider(names_from=animal, values_from=value) |>
  filter(year > 1970)
```

It's the same data, just formatted the other way around. But if we want to chart both `Sheep` and `Goats` we need the counts to be in a single column for `y`.

## Aesthetics

Aesthetics are features of the plot that allow us to distinguish one observation from another.

The most common aesthetics are:

- positions `x`, `y`
- colours `colour`, `fill`
- point style `shape`, `size`
- line style `linewidth`, `linetype`
- transparency `alpha`
- grouping `group`

## Aesthetics

:::: {.columns}

::: {.column width="50%"}
```{r gg11}
#| eval: false
ag_production |>
  filter(animal != "Goats") |>
  ggplot() +
  aes(x=year, y=value, shape=animal) +
  geom_point()
```

`shape` can be useful when we have a relatively small number of groups.

By default it is limited to 6 groups only.
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "gg11"
```
:::

::::


## Aesthetics

:::: {.columns}

::: {.column width="50%"}
```{r gg12}
#| eval: false
ggplot(data=ag_production) +
  aes(x=year, y=value, size=animal) +
  geom_point()
```

`size` is typically a continuous measure rather than a discrete one.
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "gg12"
```
:::

::::


## Aesthetics

:::: {.columns}

::: {.column width="50%"}
```{r gg13}
#| eval: false
ag_production |>
  filter(animal != "Goats") |>
  ggplot() +
  aes(x=year, y=value,
      colour=animal,
      shape=animal) +
  geom_point()
```

You can combine aesthetics and the guides will try and combine as well.
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "gg13"
```
:::

::::


## Aesthetics

:::: {.columns}

::: {.column width="50%"}
```{r gg14}
#| eval: false
ggplot(data=ag_production) +
  aes(x=year, y=value,
      linewidth=animal) +
  geom_line()
```

Line geometries can be distinguished using width and line type.
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "gg14"
```
:::

::::

## Aesthetics

:::: {.columns}

::: {.column width="50%"}
```{r gg15}
#| eval: false
ggplot(data=ag_production) +
  aes(x=year, y=value,
      linetype=animal) +
  geom_line()
```

Line geometries can be distinguished using width and line type.
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "gg15"
```
:::

::::

## Setting vs mapping an aesthetic

::: {.fragment}
If we use an aesthetic to differentiate observations, we do so by **mapping** it to a column of data via `aes()`:

```{r}
#| eval: false
ggplot(data) + aes(x=x, y=y, colour=variable) + geom_point()
```

In this case the colour of points will vary according to `variable`.

:::

::: {.fragment}
We can also **set** the aesthetic properties of the geometries we include to be the same for all observations by specifying them as parameters to the `geom_*()` functions:

```{r}
#| eval: false
ggplot(data) + aes(x=x, y=y) + geom_point(colour='pink')
```

In this case all points will be pink.

:::

## Setting vs mapping

:::: {.columns}

::: {.column width="50%"}
```{r set1}
#| eval: false
ggplot(data=ag_production) +
  aes(x=year, y=value,
      colour=animal) +
  geom_line(linewidth=2)
```

`colour` is **mapped** to `animal`.

`linewidth` is **set** to 2.
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "set1"
```
:::

::::

## Setting vs mapping

:::: {.columns}

::: {.column width="50%"}
```{r set2}
#| eval: false
ggplot(data=ag_production) +
  aes(x=year, y=value,
      colour=animal) +
  geom_point(size=3,
             alpha=0.5)
```

`colour` is **mapped** to `animal`.

`size` is set to 3.

`alpha` is set to 0.5 (half transparent).
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| ref.label: "set2"
```
:::

::::

## Geometries and Statistics

We've already seen two geometries: the point and line. ggplot2 includes about 50 or so.

Many of them are refinements on others.

- `geom_bar()` is the same as `geom_col()` but `bar` counts rows to compute `y`, whereas `col` assumes you have `y` already.

- `geom_histogram()` is just a `geom_col()` as well, where the `y` aesthetic is computed again by counting (divide `x` range into bins, count within each bin).

::: {.fragment}

The difference between these is the **statistic** used.

- `geom_col()` doesn't do anything with your data (the identity function).
- `geom_bar()` first uses `stat_count()` to count things thus producing the `y`.
- `geom_histogram()` uses `stat_bin()` to bin the `x` variable and then count within each bin.

:::

## Geometries and Statistics

Each geometry has a default **statistic** function `stat_*()`.

When your data are passed to a geometry, first the statistic is run, producing a new version of the data with additional columns for the computed statistics, along with updated aesthetic mappings.

This is then passed to one of the base geometries (points, lines, bars etc)

::: {.fragment}

```{r}
geom_histogram
```

:::

## Geometries and Statistics

Sometimes it's useful to utilise a different statistic, or a different computed value from a statistic in our charts.

For example, `stat_bin()` for histograms computes `count` (the default `y`) as well as `density`, the probability of `x` being in a particular bin.

We can override the default computed aesthetic `y` by using the `after_stat()` function.

## Count histogram

:::: {.columns}

::: {.column width="50%"}

```{r gg16}
#| eval: false
ag_production |>
  filter(animal == "Sheep") |>
  ggplot() +
  aes(x=value) +
  geom_histogram()
```

The default histogram assigns `y` to the computed variable `count`.
:::

::: {.column width="50%"}

```{r}
#| echo: false
#| ref.label: 'gg16'
```

:::

::::

## Density histogram

:::: {.columns}

::: {.column width="50%"}

```{r gg17}
#| eval: false
ag_production |>
  filter(animal == "Sheep") |>
  ggplot() +
  aes(x=value, y=after_stat(density)) +
  geom_histogram()
```

We can choose a different computed value for `y` by using `after_stat()`.

`after_stat()` implies that this value isn't available until after the statistic function has run.
:::

::: {.column width="50%"}

```{r}
#| echo: false
#| ref.label: 'gg17'
```

:::

::::

# Data wrangling

## Data wrangling

Recall that there are 5 key 'verbs' for wrangling data:

- `filter()` for choosing rows/observations
- `select()` for choosing columns/variables
- `arrange()` for ordering rows/observations
- `mutate()` for adding columns/variables
- `summarise()` for computing summaries across multiple rows/observations

## The pipe

Recall that the pipe `|>` is pronounced "and then". We use it like this:

```{r}
#| eval: false
object |> function(extra, parameters)
```

It takes the `object` on the left and inserts it as the first parameter to the `function()` on the right.

::: {.fragment fragment-index=1}

It is completely equivalent[^1] to:

```{r}
#| eval: false
function(object, extra, parameters)
```

:::

::: {.fragment fragment-index=2}

The `dplyr` verbs all take a `data.frame` or `tibble` as their first argument and return `data.frame` or `tibble`[^2]. That way we can chain them together like this:

```{r}
#| eval: false
ag_production |> filter(animal=="Sheep") |> arrange(desc(value))
```

:::

[^1]: [This is done in the R parser, so no extra code is being run here to achieve this - both expressions are parsed identically.]{.fragment fragment-index=1}

[^2]: [A `tibble` is a modern update of the `data.frame`.]{.fragment fragment-index=2}

## `filter()`

The `filter()` argument takes a boolean *expression* that returns true or false for each row in the dataset.

The expression might be "is it equal to" using `==` or "is it greater than" with `>`. Or it might be something more complicated.

```{r}
ag_production |> filter(str_detect(animal, "Cattle"))
```

## `arrange()`

We order rows with `arrange()`.

Typically we'd pass it a column to order by.

We can change to descending using the `desc()` helper:

```{r}
ag_production |> arrange(desc(value))
```

## Chaining together with the pipe

To find the year with the most pigs:

```{r}
ag_production |> filter(animal == "Pigs") |>
  arrange(desc(value))
```

## `select()`

Select is used for choosing columns. The uses are mainly:

- Presenting data in a table.
- "Eye balling" data during interactive processing (e.g. what is in column 11?).
- Janitor work on the data (dropping columns we don't care about).

```{r}
ag_production |> filter(animal == "Poultry") |>
  select(year, value)
```

## `mutate()`

We use `mutate()` to create new columns, often using existing columns in a computation.

```{r}
ag_production |> mutate(millions = round(value/1e6, 1))
```

## `mutate()`

Often once we've used columns we might want to drop them.

:::: {.columns}

::: {.column width="50%"}

```{r}
ag_production |>
  mutate(millions = round(value/1e6, 1)) |>
  select(-value)
```
:::

::: {.column width="50%"}

```{r}
ag_production |>
  mutate(millions = round(value/1e6, 1),
         .keep = 'unused')
```
:::

::::

## `summarise()`

We can `summarise()` the data by applying a summary function to all rows.

```{r}
ag_production |> summarise(max_animals = max(value),
                           min_animals = min(value))
```

::: {.fragment}

This summary isn't particularly useful though, as the `max_animals` will be from `Sheep` whereas the `min_animals` will be from something else.

:::

::: {.fragment}

A more useful summary might be for just sheep:

```{r}
ag_production |> filter(animal == "Sheep") |>
  summarise(max_sheep = max(value),
            min_sheep = min(value))
```

:::

## Grouping

Even more useful would be to find the maximum and minimum number of animals for all animal types. As our data are tidy, we can do this by grouping by animal:

```{r}
ag_production |>
  group_by(animal) |>
  summarise(min_animals = min(value),
            max_animals = max(value))
```

::: aside
A shortcut for 'summarise by group' behaviour is `ag_production |> summarise(min_animals = min(value), .by=animal)`
:::

## Joining datasets

What if we wanted to address the humans versus sheep question?

We need to take our `ag_production` data and join it up with our `population` data.

The most common join operation is

```{r}
#| eval: false
data1 |> left_join(data2, by=join_by(key1, key2, ...))
```

where columns from a second dataset are added to the data on the left.

- All rows of the left dataset are returned.

- Rows are found in the second dataset based on matches in the `key` columns.

- Columns (excluding the keys) from the second dataset are then added in.

- If the second dataset doesn't have a match it'll be given an `NA` for those columns.

- If the second dataset has multiple matches, then the corresponding rows are duplicated.

## Joining datasets

We join these datasets using `year` as the key:

```{r}
combined <- ag_production |> left_join(population, by=join_by(year))
combined
```

## Joining datasets

Notice that the combined data has `NA` values for years where there wasn't a match,and has repeated the `population` data across rows with the same `year` match:

```{r}
combined |> filter(year %in% c(1936, 1971))
```

Similarly, if `population` had two entries for 1971, they would both appear in the resulting data for each row in `ag_production`, resulting in 8 rows (4 lots of 2).

## Sheep vs Humans

:::: {.columns}

::: {.column width="50%"}

```{r sheep_vs_humans}
#| eval: false
combined |>
  filter(animal == "Sheep",
         !is.na(population)) |>
  mutate(ratio = value/population) |>
  ggplot() +
  aes(x=year, y=ratio) +
  geom_line()
```

:::

::: {.column width="50%"}

```{r}
#| echo: false
#| ref.label: "sheep_vs_humans"
```

:::

::::

## Beast vs Human

:::: {.columns}

::: {.column width="50%"}

```{r beast_vs_humans}
#| eval: false
combined |>
  filter(animal != "Sheep",
         !is.na(population)) |>
  mutate(ratio = value/population) |>
  ggplot() +
  aes(x=year, y=ratio) +
  geom_line() +
  facet_wrap(vars(animal))
```

:::

::: {.column width="50%"}

```{r}
#| echo: false
#| ref.label: "beast_vs_humans"
```

:::

::::
