---
title: "Scales, coordinates and amounts"
format:
  revealjs:
    theme: [solarized, slides.scss]
    width: 1280
    height: 720
    margin: 0.1
    smaller: false
    echo: false
    message: false
    fig-width: 7
    fig-height: 7
---

# Coordinates and scales

```{r}
library(tidyverse)
theme_set(theme_minimal(base_size=20))

ag_production <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/nz/ag_production.csv")
arxiv <- read_csv(here::here("data/arxiv_submissions.csv"))
```

## Most charts use Cartesian coordinates

```{r}
#| fig.width: 8
#| fig.height: 6
#| fig.align: 'center'
d <- tibble(x = c(-2,0,1), y=c(-1,0,1)) |>
  mutate(label = paste0('(',x,',',y,')'), nudge_x = if_else(x > 0, 0.1, -0.1), nudge_y = if_else(y < 0, -0.1, 0.1),
         hjust = if_else(x > 0, 0, 1), vjust = if_else(y < 0, 1, 0))
ggplot(d) +
  aes(x=x, y=y, label=label) +
  geom_point(col='royalblue', size=4) +
  geom_text(aes(nudge_x=nudge_x, nudge_y=nudge_y,
                hjust=hjust, vjust=vjust), size=18/.pt) +
  theme_minimal(base_size=20) +
  theme(panel.grid.minor = element_blank()) +
  coord_fixed(xlim=c(-3,2), ylim=c(-2,2)) +
  labs(x = 'x axis', y = 'y axis')
```

## Units don't matter in cartesian coordinates

::::{.columns}

:::{.column}
```{r}
ag_production |>
  filter(animal == "Sheep") |>
  ggplot() +
  aes(x=year, y=value) +
  geom_line(colour='royalblue') +
  labs(y = "Number of sheep") +
  theme_minimal(base_size=20)
```
:::

:::{.column}

:::{.fragment}
```{r}
sheep_chart <- ag_production |>
  filter(animal == "Sheep") |>
  mutate(value = value/1e6) |>
  ggplot() +
  aes(x=year, y=value) +
  geom_line(colour='royalblue') +
  labs(y = "Millions of sheep") +
  theme_minimal(base_size=20)
sheep_chart
```
:::

:::

::::

## Aspect ratio may not matter either

```{r}
#| fig.width: 10
#| fig.height: 7
#| fig.align: 'center'
library(patchwork)
(sheep_chart + sheep_chart + plot_layout(widths=c(1,2))) / sheep_chart  & theme_minimal(base_size=16)
```

## Logarithmic scales

Some measures change **multiplicatively** rather than additively.

:::{.fragment}
You start with just two rabbits, but soon you've got 4, then 8, then 16...
:::

:::{.fragment}
There's a constant increase but it is done through multiplying.
:::

## Visualising doubling 1, 2, 4, 8, 16{.smaller}

::: {.fragment style="text-align: center;"}
Original data, linear scale
```{r}
#| fig.height: 1
#| fig.width: 8
#| fig.align: 'center'
d <- tibble(x=2^(0:4))
ggplot(d) +
  aes(x=x) +
  geom_point(y=0, col='royalblue', size=4) +
  scale_y_continuous(limits=c(-0.1,0.1), breaks=0, labels="") +
  scale_x_continuous(limits=c(0,16)) +
  theme_minimal(base_size=16)
```
:::

::: {.r-stack}

::: {.fragment .fade-in-then-out style="text-align: center;"}
Log base 2 data, linear scale
```{r}
#| fig.height: 1
#| fig.width: 8
d <- tibble(x=2^(0:4))
ggplot(d) +
  aes(x=log2(x)) +
  geom_point(y=0, col='royalblue', size=4) +
  scale_y_continuous(limits=c(-0.1,0.1), breaks=0, labels="") +
  scale_x_continuous(limits=c(0,4)) +
  theme_minimal(base_size=16) +
  labs(x = expression(log[2](x)))
```
:::

:::{.fragment style="text-align: center;"}
Log base 10 data, linear scale
```{r}
#| fig.height: 1
#| fig.width: 8
d <- tibble(x=2^(0:4))
ggplot(d) +
  aes(x=log10(x)) +
  geom_point(y=0, col='royalblue', size=4) +
  scale_y_continuous(limits=c(-0.1,0.1), breaks=0, labels="") +
  scale_x_continuous(limits=c(0,log10(16))) +
  theme_minimal(base_size=16) +
  labs(x = expression(log[10](x)))
```
:::

:::

:::{.fragment style="text-align: center;"}
Original data, log scale
```{r}
#| fig.height: 1
#| fig.width: 8
#| fig.align: 'center'
d <- tibble(x=2^(0:4))
ggplot(d) +
  aes(x=x) +
  geom_point(y=0, col='royalblue', size=4) +
  scale_y_continuous(limits=c(-0.1,0.1), breaks=0, labels="") +
  scale_x_log10(limits=c(1,16), breaks=c(1,2,4,8,16)) +
  theme_minimal(base_size=16)
```
:::

## Log scales

```{r}
biorxiv <- arxiv |> filter(archive == "bioRxiv") |> filter(count > 0)
```

```{r}
#| echo: true
#| fig-width: 10
#| fig-height: 6
ggplot(biorxiv) + aes(x=date, y=count) + geom_line()
```


## Log scales

```{r}
#| echo: true
#| fig-width: 10
#| fig-height: 6
ggplot(biorxiv) + aes(x=date, y=count) + geom_line() + scale_y_log10()
```

##  Transformations

The log10 scale is implemented through the `transform` parameter of `scale_*_continuous()`.

:::{.fragment}

```{r}
#| echo: true
#| eval: false
plot + scale_y_log10()
```

is equivalent to:

:::

:::{.fragment}
```{r}
#| eval: false
#| echo: true
plot + scale_y_continuous(transform=scales::log10_trans())
```
:::

## Transformations

The `transform` functions provide all the details needed for the transformation:

:::{.incremental}
- The function to transform from x.

- The inverse to transform back to x.

- The breaks that should be used (or a function for them) on an axes.

- The domain (i.e. valid x values) that the transformation operates on.
:::

## Scale customisation

:::: {.columns}

::: {.column}
```{r bio1}
#| eval: false
#| echo: true
ggplot(biorxiv) +
  aes(x=date, y=count) +
  geom_line()
```

The bioRxiv example.
:::

::: {.column .absolute top=0 right=-5%}

```{r}
#| ref.label: "bio1"
```
:::

::::

## Scale customisation

:::: {.columns}

::: {.column}

```{r bio2}
#| eval: false
#| echo: true
ggplot(biorxiv) +
  aes(x=date, y=count) +
  geom_line() +
  scale_x_date() +
  scale_y_continuous()
```

Add `scale` functions.

No change in figure as using defaults.
:::

::: {.column .absolute top=0 right=-5%}

```{r}
#| ref.label: "bio2"
```
:::

::::

## Scale customisation

:::: {.columns}

::: {.column}

```{r bio3}
#| eval: false
#| echo: true
ggplot(biorxiv) +
  aes(x=date, y=count) +
  geom_line() +
  scale_x_date(
    name = NULL # none!
    ) +
  scale_y_continuous(
    name = "Preprints at bioRxiv"
  )
```

The `name` parameter is another
way to do `labs(x=,y=)`.
:::

::: {.column .absolute top=0 right=-5%}

```{r}
#| ref.label: "bio3"
```
:::

::::

## Scale customisation

:::: {.columns}

::: {.column}

```{r bio4}
#| echo: true
#| eval: false
ggplot(biorxiv) +
  aes(x=date, y=count) +
  geom_line() +
  scale_x_date(
    name = NULL, # none!
    
    ) +
  scale_y_continuous(
    name = "Preprints at bioRxiv",
    limits = c(0,1600)
  )
```

The `limits` parameter changes axis limits.

Alternatives are `xlim()` and `ylim()`.
:::

::: {.column .absolute top=0 right=-5%}

```{r}
#| ref.label: "bio4"
```
:::

::::

## Scale customisation

:::: {.columns}

::: {.column}

```{r bio5}
#| echo: true
#| eval: false
ggplot(biorxiv) +
  aes(x=date, y=count) +
  geom_line() +
  scale_x_date(
    name = NULL # none!
    ) +
  scale_y_continuous(
    name = "Preprints at bioRxiv",
    limits = c(0,1600),
    breaks = seq(0, 1600, by=400)
  )
```

The `breaks` parameter sets major breaks.
:::

::: aside
`seq(a, b, by=c)` produces a sequence
from `a` to `b` with steps of size `c`.
:::

::: {.column .absolute top=0 right=-5%}

```{r}
#| ref.label: "bio5"
```
:::

::::

## Scale customisation

:::: {.columns}

::: {.column}

```{r bio6}
#| echo: true
#| eval: false
ggplot(biorxiv) +
  aes(x=date, y=count) +
  geom_line() +
  scale_x_date(
    name = NULL # none!
    ) +
  scale_y_continuous(
    name = "Preprints at bioRxiv",
    limits = c(0,1600),
    breaks = seq(0, 1600, by=400),
    labels = c("0/m", "400/m", "800/m",
               "1,200/m", "1,600/m")
  )
```

The `labels` parameter sets the labels for those breaks.
:::

::: {.column .absolute top=0 right=-5%}

```{r}
#| ref.label: "bio6"
```
:::

::::

## Scale customisation

:::: {.columns}

::: {.column}

```{r bio7}
#| echo: true
#| eval: false
ggplot(biorxiv) +
  aes(x=date, y=count) +
  geom_line() +
  scale_x_date(
    name = NULL # none!
    ) +
  scale_y_continuous(
    name = "Preprints at bioRxiv",
    limits = c(0,1600),
    breaks = seq(0, 1600, by=400),
    labels = function(x) paste0(x,"/m")
  )
```

The `labels` parameter can be a function.
:::

::: {.column .absolute top=0 right=-5%}

```{r}
#| ref.label: "bio7"
```
:::

::::

## Scale customisation

:::: {.columns}

::: {.column}

```{r bio8}
#| echo: true
#| eval: false
ggplot(biorxiv) +
  aes(x=date, y=count) +
  geom_line() +
  scale_x_date(
    name = NULL # none!
    ) +
  scale_y_continuous(
    name = "Preprints at bioRxiv",
    limits = c(0,1600),
    breaks = seq(0, 1600, by=400),
    labels = scales::label_comma(
      suffix="/m"
      )
  )
```

The `scales` package has lots of common labellers.
:::

::: {.column .absolute top=0 right=-5%}

```{r}
#| ref.label: "bio8"
```
:::

::::

## Scale customisation

:::: {.columns}

::: {.column}

```{r bio9}
#| echo: true
#| eval: false
ggplot(biorxiv) +
  aes(x=date, y=count) +
  geom_line() +
  scale_x_date(
    name = NULL, # none!
    date_breaks = "2 years"
    ) +
  scale_y_continuous(
    name = "Preprints at bioRxiv",
    limits = c(0,1600),
    breaks = c(0, 400, 800, 1200, 1600),
    labels = scales::label_comma(
      suffix="/m"
      )
  )
```

Some scales have alternate versions of breaks.
:::

::: {.column .absolute top=0 right=-5%}

```{r}
#| ref.label: "bio9"
```
:::

::::

## Polar coordinates

The **coordinate** system defaults to cartesian.

Polar coordinates are suited where data are 'circular' in some way.

:::{.incremental}
- hourly or within-day patterns

- daily or within week patterns

- yearly patterns (seasonality)

- longitude (maps!)
:::

## Solar generation in Palmerston North

```{r}
solar <- read_csv(here::here("data/solar_by_day.csv")) |>
  mutate(kWh = smoothed/1e3)
```

```{r}
#| echo: true
#| fig-width: 10
#| fig-height: 7
ggplot(solar) + aes(x=day, y=kWh) + geom_line()
```

## Solar in Palmerston North

```{r}
#| echo: true
#| fig-width: 10
#| fig-height: 7
ggplot(solar) + aes(x=day, y=kWh) + geom_line() + coord_polar()
```

## Fixing up the kWh scale

```{r}
#| echo: true
#| fig-width: 10
#| fig-height: 7
ggplot(solar) + aes(x=day, y=kWh) + geom_line() + coord_polar() +
  scale_y_continuous(limits=c(0,40))
```

# Visualising Amounts

## Amounts

Often we have data that are as simple as some amounts.

:::{.fragment}

:::: {.columns}

::: {.column}
**Example:** Millions of production animals in 2025.
:::

:::{.column}
```{r}
animals <- 
  ag_production |>
  slice_max(year) |> select(-year) |>
  mutate(value = value/1e6)

knitr::kable(animals, digits=1)
```
:::

::::

:::

## Bar charts are usually best

```{r}
#| fig.width: 10
#| fig.height: 6
#| fig.align: 'center'
ggplot(animals) +
  aes(x=fct_rev(fct_reorder(animal, value)), y=value) +
  geom_col(width=0.6, fill='steelblue') +
  scale_y_continuous(expand=expansion(mult = c(0,0.05))) +
  labs(x=NULL, y="Number of animals (m)") +
  theme(axis.ticks.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank())
```

## Long labels on angles?

```{r}
#| fig.width: 10
#| fig.height: 6
#| fig.align: 'center'
ggplot(animals) +
  aes(x=fct_rev(fct_reorder(animal, value)), y=value) +
  geom_col(width=0.6, fill='steelblue') +
  scale_y_continuous(expand=expansion(mult = c(0,0.05))) +
  labs(x=NULL, y="Number of animals (m)") +
  annotate('text', x=6.5, y=25, label="yeah nah", col='darkred', size=10) +
  theme(axis.ticks.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 30, hjust=1))
```

## We have the power of horizontal bars

```{r}
#| fig.width: 10
#| fig.height: 6
#| fig.align: 'center'
ggplot(animals) +
  aes(y=fct_reorder(animal, value), x=value) +
  geom_col(width=0.7, fill='steelblue') +
  scale_x_continuous(expand=expansion(mult = c(0,0.05))) +
  labs(y=NULL, x="Number of animals (m)") +
  annotate('text', x=20, y=1, label="yeah nah yeah", col='green4', size=10) +
  theme(axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank())
```

## Points can also be used

```{r}
#| fig.width: 10
#| fig.height: 6
#| fig.align: 'center'
lego <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/lego/lego_sets.csv")
sets_per_year <- lego |>
  count(theme, year) |>
  summarise(sets=median(n), .by=theme) |>
  filter(sets > 5)
ggplot(sets_per_year) +
  aes(y=fct_reorder(theme, sets), x=sets) +
  geom_point(col='steelblue', size=5) +
  scale_x_continuous(limits=c(0, 51), expand=expansion(mult = c(0,0.05))) +
  labs(y=NULL, x="Median sets per year") +
#  annotate('text', x=20, y=1, label="yeah nah yeah", col='green4', size=10) +
  theme(axis.ticks.y = element_blank(),
        panel.grid.minor = element_blank())
```

## Bars should start at zero

```{r}
#| fig.width: 10
#| fig.height: 6
#| fig.align: 'center'
lego <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/lego/lego_sets.csv")
sets_per_year <- lego |>
  count(theme, year) |>
  summarise(sets=median(n), .by=theme) |>
  filter(sets > 5)
ggplot(sets_per_year) +
  aes(y=fct_reorder(theme, sets), x=sets) +
  geom_col(fill='steelblue', width=0.6) +
  scale_x_continuous(limits=c(0, 51), expand=expansion(mult = c(0,0.05))) +
  coord_cartesian(xlim=c(10, 51)) +
  labs(y=NULL, x="Median sets per year") +
  annotate('text', x=45, y=1, label="yeah nah", col='darkred', size=10) +
  theme(axis.ticks.y = element_blank(),
        panel.grid.minor = element_blank())
```

## Use points if zero is less important

```{r}
#| fig.width: 10
#| fig.height: 6
#| fig.align: 'center'
ggplot(sets_per_year) +
  aes(y=fct_reorder(theme, sets), x=sets) +
  geom_point(col='steelblue', size=5) +
  scale_x_continuous(limits=c(10, 51), expand=expansion(mult = c(0,0.05))) +
  labs(y=NULL, x="Median sets per year") +
  annotate('text', x=45, y=1, label="yeah nah yeah", col='green4', size=10) +
  theme(axis.ticks.y = element_blank(),
        panel.grid.minor = element_blank())
```

## Choose the 'right' order

:::: {.columns}

::: {.column}
```{r order1}
#| eval: false
#| echo: true
ggplot(sets_per_year) +
  aes(y=theme,
      x=sets) +
  geom_col()
```

The default order is alphabetical.

The `y` axis points up.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'order1'
```
:::

::::

## Choose the 'right' order

:::: {.columns}

::: {.column}
```{r order2}
#| eval: false
#| echo: true
ggplot(sets_per_year) +
  aes(y=fct_reorder(theme, sets),
      x=sets) +
  geom_col()
```

`fct_reorder` will reorder the
grouping variable by the measure.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'order2'
```
:::

::::


## Choose the 'right' order

:::: {.columns}

::: {.column}
```{r order3}
#| eval: false
#| echo: true
ggplot(sets_per_year) +
  aes(x=fct_reorder(theme, sets),
      y=sets) +
  geom_col()
```

Ascending order not as good for `x`.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'order3'
```
:::

::::

## Choose the 'right' order

:::: {.columns}

::: {.column}
```{r order4}
#| eval: false
#| echo: true
ggplot(sets_per_year) +
  aes(x=fct_rev(
          fct_reorder(theme, sets)
        ),
      y=sets) +
  geom_col()
```

`fct_rev` to reverse.

Alternatively `fct_reorder(.desc=TRUE)`.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'order4'
```
:::

::::

## Grouped bars

```{r}
set_counts <- lego |>
  count(theme, year, name = 'sets') |>
  filter(year > 2015, year < 2021) |>
  group_by(theme) |>
  filter(sum(sets) > 30) |>
  mutate(year = as_factor(year))
options('ggplot2.discrete.fill' = scales::viridis_pal(option="rocket")(6))
```

:::: {.columns}

::: {.column}
```{r group1}
#| eval: false
#| echo: true
ggplot(set_counts) +
  aes(x=theme,
      y=sets,
      fill=year) +
  geom_col()
```

By default groups are stacked.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'group1'
```
:::

::::

## Grouped bars


:::: {.columns}

::: {.column}
```{r group2}
#| eval: false
#| echo: true
ggplot(set_counts) +
  aes(x=theme,
      y=sets,
      fill=year) +
  geom_col(position='dodge')
```

Side by side is often better.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'group2'
```
:::

::::


## Grouped bars

:::: {.columns}

::: {.column}
```{r group3}
#| eval: false
#| echo: true
ggplot(set_counts) +
  aes(x=year,
      y=sets,
      fill=theme) +
  geom_col(position='dodge')
```

Can swap `fill` and `x`.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'group3'
```
:::

::::

## Grouped bars

:::: {.columns}

::: {.column}
```{r group4}
#| eval: false
#| echo: true
ggplot(set_counts) +
  aes(x=year,
      y=sets) +
  geom_col() +
  facet_wrap(vars(theme))
```

Facetting is another option.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'group4'
```
:::

::::

## Counting first?

We used `geom_col()` for LEGO sets as our data was already counted.

::: {.fragment}
If your data aren't summarised, then you can:

::: {.incremental}
- Summarise it first with `dplyr::count()`.

- Use `geom_bar()` so that `stat_count()` does it for you.
:::
:::

## Counting first?

Consider the uncounted lego data for 2020:

```{r}
uncounted <- lego |> filter(year == 2020) |>
  add_count(theme) |> filter(n > 6) |> select(set, name, theme)
uncounted
```

## Counting first?

:::: {.columns}

::: {.column}
```{r count1}
#| eval: false
#| echo: true
ggplot(uncounted) +
  aes(y=theme) +
  geom_bar()
```

`geom_bar` uses `stat_count()` to count each `theme` to get `x`.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'count1'
```
:::

::::

## Counting first?

:::: {.columns}

::: {.column}
```{r count2}
#| eval: false
#| echo: true
ggplot(uncounted) +
  aes(y=fct_infreq(theme)) +
  geom_bar()
```

`fct_infreq()` will order by frequency.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'count2'
```
:::

::::

## Counting first?

:::: {.columns}

::: {.column}
```{r count2b}
#| eval: false
#| echo: true
ggplot(uncounted) +
  aes(y=fct_rev(
          fct_infreq(theme)
        )) +
  geom_bar()
```

Reversing is required for `y`.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'count2b'
```
:::

::::

## Counting first?

:::: {.columns}

::: {.column}
```{r count3}
#| eval: false
#| echo: true
uncounted |>
  count(theme) |>
  ggplot() +
  aes(y=theme, x=n) +
  geom_col()
```

Alternatively we can count first and use `geom_col()`
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'count3'
```
:::

::::

## Counting first?

:::: {.columns}

::: {.column}
```{r count4}
#| eval: false
#| echo: true
uncounted |>
  count(theme) |>
  ggplot() +
  aes(y=fct_reorder(theme, n),
      x=n) +
  geom_col()
```

And use `fct_reorder()` for the count order.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'count4'
```
:::

::::

# Proportions

## Proportions

The traditional chart for proportions of a whole is the pie.

::: {.fragment}
The problems with a pie are two fold:

::: {.incremental}
- Humans are only good at large differences or simple fractions (1/2, 1/4).

- "Hard" to do in `ggplot2`: use a bar with polar coordinates.
:::

:::

::: {.fragment}
Often a barchart (stacked or unstacked) is just as good. But not always!
:::

## Pie charts in ggplot2

```{r}
sets_2020 <- set_counts |> filter(year == 2020) |>
  ungroup() |>
  mutate(theme = fct_reorder(theme, sets))
```

:::: {.columns}

::: {.column}

```{r pie1}
#| eval: false
#| echo: true
ggplot(sets_2020) +
  aes(x=sets, y = "YY",
      fill=theme) +
  geom_col()
```

Start with a stacked bar chart.

Y axis is set to a single value so that `fill` stacks.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'pie1'
```
:::

::::

## Pie charts in ggplot2

:::: {.columns}

::: {.column}

```{r pie2}
#| eval: false
#| echo: true
ggplot(sets_2020) +
  aes(x=sets, y = "YY",
      fill=theme) +
  geom_col() +
  coord_polar()
```

Switch to polar coordinates.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'pie2'
```
:::

::::


## Pie charts in ggplot2

:::: {.columns}

::: {.column}

```{r pie3}
#| eval: false
#| echo: true
ggplot(sets_2020) +
  aes(x=sets, y = "YY",
      fill=theme) +
  geom_col() +
  coord_polar() +
  scale_x_continuous(name=NULL,
                   breaks=NULL) +
  scale_y_discrete(name=NULL,
                   breaks=NULL)
```

Drop the axis names and breaks.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'pie3'
```
:::

::::


## Pie charts in ggplot2

:::: {.columns}

::: {.column}

```{r pie4}
#| eval: false
#| echo: true
ggplot(sets_2020) +
  aes(x=sets, y = "YY",
      fill=theme) +
  geom_col() +
  coord_polar() +
  scale_x_continuous(name=NULL,
                   breaks=NULL) +
  scale_y_discrete(name=NULL,
                   breaks=NULL) +
  theme(panel.grid=element_blank())
```

Remove panel grid.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'pie4'
```
:::

::::


## Pie vs bar

```{r}
#| echo: false
#| fig.width: 12
#| fig.height: 7
#| fig.align: 'center'
library(patchwork)
pie <- ggplot(sets_2020) +
  aes(x=sets, y = "YY",
      fill=theme) +
  geom_col() +
  coord_polar() +
  scale_x_continuous(name=NULL,
                   breaks=NULL) +
  scale_y_discrete(name=NULL,
                   breaks=NULL) +
  theme(panel.grid=element_blank())

bar <- ggplot(sets_2020) +
  aes(x=sets, y=theme, fill=theme) +
  geom_col() +
  scale_x_continuous(expand=expansion(mult=c(0,0.05))) +
  guides(fill='none') +
  theme(panel.grid.major.y=element_blank(),
        panel.grid.minor=element_blank()) +
  labs(y=NULL, x = "Number of sets")

pie + bar
```

## Making bars prettier

:::: {.columns}

::: {.column}

```{r pretty1}
#| eval: false
#| echo: true
ggplot(sets_2020) +
  aes(x=sets, y=theme) +
  geom_col()
```

Start simple.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'pretty1'
```
:::

::::

## Making bars prettier

:::: {.columns}

::: {.column}

```{r pretty2}
#| eval: false
#| echo: true
ggplot(sets_2020) +
  aes(x=sets, y=theme) +
  geom_col() +
  scale_x_continuous(
    expand=expansion(mult=c(0,0.05))
    )
```

Change `expand` so the axis is at 0.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'pretty2'
```
:::

::::

## Making bars prettier

:::: {.columns}

::: {.column}

```{r pretty3}
#| eval: false
#| echo: true
ggplot(sets_2020) +
  aes(x=sets, y=theme, fill=theme) +
  geom_col() +
  scale_x_continuous(
    expand=expansion(mult=c(0,0.05))
    ) +
  guides(fill='none')
```

Add some colour.

Turn off the guide.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'pretty3'
```
:::

::::


## Making bars prettier

:::: {.columns}

::: {.column}

```{r pretty4}
#| eval: false
#| echo: true
ggplot(sets_2020) +
  aes(x=sets, y=theme, fill=theme) +
  geom_col() +
  scale_x_continuous(
    expand=expansion(mult=c(0,0.05))
    ) +
  guides(fill='none') +
  labs(y=NULL, x = "Number of sets")
```

Improve the labels.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'pretty4'
```
:::

::::

## Making bars prettier

:::: {.columns}

::: {.column}

```{r pretty5}
#| eval: false
#| echo: true
ggplot(sets_2020) +
  aes(x=sets, y=theme, fill=theme) +
  geom_col() +
  scale_x_continuous(
    expand=expansion(mult=c(0,0.05))
    ) +
  guides(fill='none') +
  labs(y=NULL, x = "Number of sets") +
  theme(panel.grid.major.y=element_blank(),
        panel.grid.minor=element_blank())
```

Tweak the theme to remove some of the grid.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'pretty5'
```
:::

::::


## Making bars prettier

:::: {.columns}

::: {.column}

```{r pretty6}
#| eval: false
#| echo: true
ggplot(sets_2020) +
  aes(x=sets, y=theme, fill=theme) +
  geom_col(width=0.8) +
  scale_x_continuous(
    expand=expansion(mult=c(0,0.05))
    ) +
  guides(fill='none') +
  labs(y=NULL, x = "Number of sets") +
  theme(panel.grid.major.y=element_blank(),
        panel.grid.minor=element_blank())
```

Reduce the bar width.

:::{.fragment}
More of this in week 3!
:::
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'pretty6'
```
:::

::::
