---
title: "Space!"
format:
  revealjs:
    theme: [solarized, slides.scss]
    width: 1280
    height: 720
    margin: 0.1
    echo: false
    message: false
    fig-width: 7
    fig-height: 7
---

```{r}
#| echo: false

library(tidyverse)
theme_set(theme_minimal(base_size=20))
```

# Spatial data

## Spatial data

Spatial data is anything that has some position or coordinates associated with it.

There are four main types:

- **Points**: x, y coordinates possibly with some other data associated with them.

- **Lines**: A linked set of points that may have data on the links.

- **Polygons**: closed polygons that describe the boundaries of a set of regions which possibly have some other data associated with them.

- **Grids**: A special case where the polygons are uniform rectangles that form a grid.

We'll be looking at points, polygons and grids. Lines will be present through polygon borders though!

## Points

Point data are the locations of observations through some spatial region.

The spatial region might be bounded (e.g. points within a country or city) or unbounded.

Can be more than two dimensional (e.g. points in 3D space) but we'll be focusing on 2D, and will generally assume that this 2D data are **planar**. i.e. lie in a plane.

Many such data that we assume are planar are not! e.g. locations on the planet are better approximated by a sphere but like the term "plane sailing" we'll temporally become flat-earthers!

## Questions about points

The simplest question about points is: "why are they there?"

i.e. what is the **process** that gave rise to this particular pattern of points that we see.

Other questions may be:

- Is something we've measured at these points related to the location? *Spatial correlation*

- Can we take a measure from one set of points and infer what that measure might be somewhere else? *Spatial interpolation*

## Earthquakes around Fiji

```{r}
ggplot(quakes) +
  aes(x=long, y=lat) +
  geom_point(alpha=0.2)
```

- Have longitude/latitude information
- Assume the data are 'flat' in these
- Aspect ratio will matter though!

## Cricket wagon-wheels

```{r}
nz_cric <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/cricket/nz_bat_scores.csv")

bmc <- nz_cric |> filter(p_match == 667653,
                  score != 0, inns == 3,
                  wagonX != 0, wagonY != 0,
                  bat == "Brendon McCullum")
ggplot(bmc) +
  aes(x=wagonX, y=wagonY, colour=outcome, linewidth=outcome) +
  geom_segment(aes(xend=0, yend=0)) +
  scale_colour_manual(values = c(six = 'red2', four='gold', run='white')) +
  scale_linewidth_manual(values= c(six = 1.5, four=1, run=0.5)) +
  theme_void() + 
  theme(panel.background=element_rect(fill='green4')) +
  guides(colour='none', linewidth='none') +
  coord_fixed()
```

Brendon McCullum's innings of
302 vs India in 2014

## Cricket wagon-wheels

```{r}
ggplot(bmc) +
  aes(x=wagonX, y=wagonY) +
  geom_point(alpha=0.3)
```

Brendon McCullum's innings of
302 vs India in 2014 as a point pattern.

## Point processes

A point process is the statistical model that gives rise to the location of points.

A one-dimensional point process produces event times.

A two-dimensional point process produces event locations.

We usually model these with some sort of *statistical process*.

## Poisson processes

A spatial *Poisson process* generates $N$ points in an area $A$ proportionally according to the area of the region and some underlying "points per unit area" rate, $\lambda$. That is, for any region $A$ and rate $\lambda$ we expect the number of points $N(A)$ to follow a Poisson distribution:

$$
P(N(A) = n) ~ Poisson(\lambda|A|)
$$

The rate of points scales with the area: larger area has more points than smaller areas.

The **strong** assumption is that the points don't interact: i.e. their location is not influenced by the location of other points.

## Random?

Picture of homogeneous PPT and a couple of cluster processes.

## Homogeneous and inhomogeneous processes

If the rate $\lambda$ is constant across the region, then the process is **homogeneous**, and the points that result are completely spatially random: The chance of a point being in any location is fully randomised.

In most cases the chance of a point being in a location is not fully randomised as some locations are more likely than others. e.g. people may enjoy living by the sea more than they enjoy living in a desert, so locations closer to the sea will be more likely to have people than locations in deserts.

This is an **inhomogeneous** process where the rate $\lambda = \lambda(x,y)$ depends on the location $(x,y)$. The same basic ideas apply: the number of points in a region A still follow the Poisson but now we have to account for the varying rate or **intensity** $\lambda$:

$$
P(N(A) = n) ~ Poisson(\Lambda(x))
\Lambda(x) = \int_A \lambda(x) dx
$$

## Points are still independent

In an inhomogeneous process the points still don't interact with each other: having more people nearby does not mean that additional people are likely to be there, it's just that there's some external pressure from the intensity $\lambda$ that results in more points in one place than another.

Other fancier processes that allow for **spatial clustering** are also a thing.

See 233.214 GIS and Spatial Statistics for more on this!

## What is important in a point process?

The key thing of importance is the underlying **intensity** function $\lambda(x)$: we want to know where points are more or less likely to be.

We note that the data we have to do this is finite: just a pattern of $n$ points. And the thing we want to know about is an infinite dimensional object: a (probably) smooth function of x for the infinite possibilities of locations.

So we're not just estimating a mean here! We're gonna need something a bit fancier.

## Estimating point intensity

Typically we use a kernel density estimator to estimate $\lambda$, by noting that we can write $\lambda(x)$ as a constant (the expected number of points) times a probability density (i.e. sums to 1):

$$
\lambda(x) = \Lambda_0 f(x), \qquad \int_A f(x) dx = 1.
$$

Estimating $f(x)$ is then just estimating a density, just we're in two dimensions instead of one:

$$
\hat f(\mathbf{x})= \frac{1}{n} \sum_{i=1}^n K_h(\mathbf{x} - \mathbf{x}_i)
$$
Same idea, just that now $\mathbf{x}$ and $\mathbf{x}_i$ are 2-d spatial locations, and the kernel $K_h$ is a two dimensional kernel rather than 1d. We usually settle for a bivariate normal distribution and assume that it is rotationally symmetric[^1] so that $h$ is a scalar.

In `ggplot2` we can use `geom_density_2d()` :)

[^1]: isotropic

## Quakes

EXAMPLE GOES HERE

## Cricket

EXAMPLE GOES HERE

## KDEs give uniform smoothing

Our kernel density estimator uses a single bandwidth $h$ for the smoothing.

This applies across the entire area.

But, if the intensity is changing across the area, then that implies that the amount of data (information!) we have differs across the region.

If we smooth everything by the same amount, places where we have lots of data/signal will be tend to be smoothed too much, whilst places where data are sparse may be smoothed too little.

Ideally we'd allow our estimator to adapt the amount of smoothing to the situation.

## Adaptive KDE

In adaptive kernel density estimates we allow $h = h(x)$ to vary with $x$.

But this presents a chicken and egg situation: the ideal is for $h(x)$ to be larger (more smoothing) when points are rare (i.e. when $f(x)$ is small) while $h(x)$ should be smaller (less smoothing) when points are dense (i.e. when $f(x)$ is large). But we don't know $f(x)$!!

## Iterating out of the chicken/egg

To figure out the best $h(x)$ we first obtain a pilot estimate $\hat f_p(x)$ with constant smoothing.

We then set $h(x) = \frac{1}{\sqrt{f_p(x)}}$ which turns out to be optimal[^2] and use that in a second round of estimation.

Potentially we could then iterate again, using our new (adaptive) estimate to get a better idea of $h(x)$ to estimate a third time, but it's usually not necessary!

[^2]: See Abramson 1982

## Adaptive cricket?


## Adaptive quakes

TODO: Constant, then show bandwidth (why 1/sqrt? - it's circular, and num points scales with area not radius) then adaptive.

## Adding context to the quakes?

e.g. Where is New Zealand in relation to this map?

A map of NZ is available via map_data('nz')

This just gives a row of x/y coordinates.

Can use geom_path to link them up (difference between path/line/polygon)

Sidequest: geom_line() orders the data in x first (good for time-series as the plot really shouldn't depend on order of x, right?) geom_path() orders by order in the data.

So draw NZ on.

## But is this flat? The sailing isn't plane anymore.

Show distance between lowest and highest latitudes.

We should at least be aware of the distortion that results (next time: CRS!)

## Further reading...

