---
title: "Maps"
format:
  revealjs:
    theme: [solarized, slides.scss]
    width: 1280
    height: 720
    margin: 0.1
    echo: true
    message: false
    fig-width: 7
    fig-height: 8
---

```{r}
#| echo: false

library(tidyverse)
theme_set(theme_minimal(base_size=20))
```

# Polygons

## Polygon maps

The simplest form of map data are polygons on lat/longitude or similar:

```{r}
nz <- map_data('nz')
nz
```

## Polygon maps

:::: columns
::: column
```{r nz1}
#| eval: false
ggplot(nz) +
  aes(x=long, y=lat) +
  geom_point()
```
:::

::: {.column .absolute top="0" right="-5%"}
```{r}
#| echo: false
#| ref.label: nz1
```
:::
::::

## Polygon maps

:::: columns
::: column
```{r nz2}
#| eval: false
ggplot(nz) +
  aes(x=long, y=lat) +
  geom_line()
```

Lines can link up points?
:::

::: {.column .absolute top="0" right="-5%"}
```{r}
#| echo: false
#| ref.label: nz2
```
:::
::::

## Polygon maps

:::: columns
:::: column
```{r nz3}
#| eval: false
ggplot(nz) +
  aes(x=long, y=lat, group=group) +
  geom_line()
```

`group` separates islands.

::: fragment
`geom_line()` arranges the data by `x`.
:::
::::

::: {.column .absolute top="0" right="-5%"}
```{r}
#| echo: false
#| ref.label: nz3
```
:::
::::

## Polygon maps

:::: columns
::: column
```{r nz4}
#| eval: false
ggplot(nz) +
  aes(x=long, y=lat, group=group) +
  geom_path()
```

`group` separates islands.

`geom_path()` directly links points.
:::

::: {.column .absolute top="0" right="-5%"}
```{r}
#| echo: false
#| ref.label: nz4
```
:::
::::

## Polygon maps

:::: columns
::: column
```{r nz5}
#| eval: false
ggplot(nz) +
  aes(x=long, y=lat, group=group) +
  geom_polygon(fill='tan')
```

`geom_polygon()` for a filled polygon.
:::

::: {.column .absolute top="0" right="-5%"}
```{r}
#| echo: false
#| ref.label: nz5
```
:::
:::::

## Polygon maps

:::: columns
::: column
```{r nz6}
#| eval: false
ggplot(nz) +
  aes(x=long, y=lat, group=group) +
  geom_polygon(fill='tan') +
  coord_sf()
```

Aspect ratio matters.

`coord_sf()` can help.
:::

::: {.column .absolute top="0" right="-5%"}
```{r}
#| echo: false
#| ref.label: nz6
```
:::
:::::

## Polygon maps

```{r}
#| echo: false
petrels <- read_csv("https://www.massey.ac.nz/~jcmarsha/227212/data/petrels.csv") |>
  rename(long = Longitude, lat = Latitude, group=Area)
```

:::: columns
::: column
```{r nz7}
#| eval: false
ggplot(nz) +
  aes(x=long, y=lat, group=group) +
  geom_polygon(fill='tan') +
  geom_point(data=petrels,
             alpha=0.3) +
  coord_sf()
```

Additional data added via `data=`.
:::

::: {.column .absolute top="0" right="-5%"}
```{r}
#| echo: false
#| ref.label: nz7
```
:::
:::::

## Polygon maps

```{r}
#| echo: false
petrels <- read_csv("https://www.massey.ac.nz/~jcmarsha/227212/data/petrels.csv")
```

:::: columns
::: column
```{r nz8}
#| eval: false
ggplot(nz) +
  geom_polygon(
    mapping = aes(x=long,
                  y=lat,
                  group=group),
    fill='tan') +
  geom_point(
    data=petrels,
    mapping=aes(x=Longitude,
                y=Latitude),
    alpha=0.3) +
  coord_sf()
```

In practice need to utilise separate `aes()` lines for separate `data`.
:::

::: {.column .absolute top="0" right="-5%"}
```{r}
#| echo: false
#| ref.label: nz8
```
:::
:::::

# Simple features

## Simple features

Simple features is a standard produced by the Open Geospatial Consortium.

It defines a common storage and access model of geographic features made using two-dimensional geometries.

Codifies things like database access, geometry types, coordinate reference systems etc.

::: fragment
The `sf` package provides most of the tools needed to work with these data.

`geom_sf()` and `coord_sf()` are then the `ggplot2` functions that work with `sf`.
:::

## Simple features data

```{r}
#| echo: false
library(sf)
nz_sf <- nz |> st_as_sf(coords = c('long', 'lat')) |>
  group_by(region) |>
  summarize(do_union = FALSE, .groups='drop') |>
  st_cast("LINESTRING") |>
  st_cast("POLYGON") |>
  mutate(region = str_replace_all(region, '\\.', ' ') |> str_trim()) |>
  tibble::remove_rownames() |>
  st_set_crs(4326)
```

```{r}
nz_sf
```

::: fragment
Each row represents a single region (island) with spatial information in a `geometry` column.
:::

## Simple features

:::: columns
::: column
```{r sf1}
#| eval: false
ggplot(data=nz_sf) +
  geom_sf()
```

Can use `geom_sf()` to draw directly.
:::

::: {.column .absolute top="0" right="-5%"}
```{r}
#| echo: false
#| ref.label: sf1
```
:::
:::::

## Simple features

:::: columns
::: column
```{r sf2}
#| eval: false
ggplot(data=nz_sf) +
  aes(geometry=geometry) +
  geom_sf()
```

Can use `geom_sf()` to draw directly.

The `geometry` aesthetic is specific to `sf` data.

Automatically detected if there is only one `geometry` column.
:::

::: {.column .absolute top="0" right="-5%"}
```{r}
#| echo: false
#| ref.label: sf2
```
:::
:::::

## Simple features

:::: columns
::: column
```{r sf3}
#| eval: false
ggplot(data=nz_sf) +
  aes(fill=region) +
  geom_sf() +
  guides(fill = 'none')
```

Other aesthetics work as expected.
:::

::: {.column .absolute top="0" right="-5%"}
```{r}
#| echo: false
#| ref.label: sf3
```
:::
:::::

## Simple features

The metadata block tells us a bit about the simple features data we're working with:

```{r}
#| echo: false
print(nz_sf, n=0)
```

- The number of features and fields (non-geometry columns)

- The geometry type and dimension

- A rectangular boundary that contains all points

- The coordinate reference system (more later!)

## Layering

:::: columns
::: column
```{r sf4}
#| eval: false
ggplot(data=nz_sf) +
  geom_sf() +
  geom_point(data=petrels,
             aes(x=Longitude,
                 y=Latitude,
                 col=Area)) +
  guides(col='none')
```

Can add other datasets and use additional geometries.
:::

::: {.column .absolute top="0" right="-5%"}
```{r}
#| echo: false
#| ref.label: sf4
```
:::
:::::

# Coordinate reference systems

## Map projections

The planet is not flat! It's not a sphere either. So when we map a coordinate (e.g. longitude/latitude) to a location we need to make lots of assumptions:

::: incremental
- How much of an ellipsoid is the Earth?

- Where is its centre?

- Where is sea level?

- Where is the origin for our x/y pairs?

- How do the techtonic plates move?
:::

::: fragment
How relevant each is depends on how far apart our data are: if they're all within a few hundred kilometers, 'plane sailing' is probably just fine!
:::

## Map projections

The set of assumptions about earth shape and it's link to coordinate pairs is known as the **geodetic datum**.

WGS84 (the world geodetic system) is the classic world datum featured on most online mapping things (Open Street Maps etc).

It defines:

- The ellipsoid

- The origin for longitude (Greenwich) and latitude (Equator)

- What a one unit increase in each means (change of angle of one degree).

## WGS84

```{r}
st_crs(nz_sf)
```

## NZTM2000

The most common NZ-wide projection is probably NZTM2000, though there are many others that focus on specific areas.

It is a Transverse Mercator (cylindrical north/south) projection which works well as NZ is largely orientated north/south.

It uses the GRS80 oblate ellipisoid (flattened at the poles) and the New Zealand Geodetic Datum 2000 (NZGD2000).

The origin for NZTM is 173°E, 0°S which are represented with easting/northing (x/y) values in metres of 1,600,000m and 100,000,000m.

## NZTM2000

```{r}
nz_sf |> st_transform(crs=2193) # NZTM2000 has ESPG code 2193
```

Now our polygon coordinates are expressed in metres rather than degrees.

## NZTM 2000

:::: columns
::: column
```{r sf5}
#| eval: false
nz_sf |>
  st_transform(crs=2193) |>
  ggplot() +
  geom_sf()
```

This reflects the cylindrical projection.
:::

::: {.column .absolute top="0" right="-5%"}
```{r}
#| echo: false
#| ref.label: sf5
```
:::
:::::

## NZTM 2000

:::: columns
::: column
```{r sf6}
#| eval: false
ggplot(nz_sf) +
  geom_sf() +
  coord_sf(crs=2193)
```

Can also do it at chart time with `coord_sf()`.
:::

::: {.column .absolute top="0" right="-5%"}
```{r}
#| echo: false
#| ref.label: sf6
```
:::
:::::

## Another projection

:::: columns
::: column
```{r sf7}
#| eval: false
ggplot(nz_sf) +
  geom_sf() +
  coord_sf(crs=9191) # Niwa Albers
```

Lots of other projections are available.
:::

::: {.column .absolute top="0" right="-5%"}
```{r}
#| echo: false
#| ref.label: sf7
```
:::
::::

## Census 2023 data

```{r}
library(sf)
nz2023 <- read_sf("https://www.massey.ac.nz/~jcmarsha/data/nz_map/sa3_data.sqlite")
nz2023
```

## Census 2023

:::: columns
::: column
```{r census1}
#| eval: false
ggplot(nz2023) +
  aes(fill=moved_in_last_year) +
  geom_sf(linewidth=0.1) +
  scale_fill_viridis_c() +
  labs(fill=NULL,
       title="Moved recently")
```

:::

::: {.column .absolute top="0" right="-5%"}
```{r}
#| echo: false
#| ref.label: census1
```
:::
::::

## Census 2023

:::: columns
::: column
```{r census2}
#| eval: false
nz2023 |>
  filter(territorial_authority == "Auckland") |>
  ggplot() +
  aes(fill=moved_in_last_year) +
  geom_sf(linewidth=0.1) +
  scale_fill_viridis_c() +
  labs(fill=NULL,
       title="Moved recently")
```

:::

::: {.column .absolute top="0" right="-5%"}
```{r}
#| echo: false
#| ref.label: census2
```
:::
::::


## Census 2023

:::: columns
::: column
```{r census3}
#| eval: false
nz2023 |>
  filter(territorial_authority == "Auckland") |>
  ggplot() +
  aes(fill=median_income) +
  geom_sf(linewidth=0.1) +
  scale_fill_viridis_c() +
  labs(fill=NULL,
       title="Median income")
```

:::

::: {.column .absolute top="0" right="-5%"}
```{r}
#| echo: false
#| ref.label: census3
```
:::
::::

## Census 2023

:::: columns
::: column
```{r census4}
#| eval: false
nz2023 |>
  filter(territorial_authority == "Auckland") |>
  ggplot() +
  aes(fill=home_for_last) +
  geom_sf(linewidth=0.1) +
  scale_fill_viridis_c() +
  labs(fill=NULL,
       title="Years in same home")
```

:::

::: {.column .absolute top="0" right="-5%"}
```{r}
#| echo: false
#| ref.label: census4
```
:::
::::

## Rasters

While simple features data are polygon, many datasets work better on a grid.

Such grids are represented as pixel images or **rasters**.

They needn't be strict grids: they might be non-equispaced, skewed or curvi-linear.

Most are grids though!

## Rasters

The `stars` and `terra` packages are for working with rasters. `stars` provides `geom_stars()` for
adding rasters to charts.

Forest coverage data for NZ from Manaaki Whenua Landcare Research at 2km resolution:

```{r}
library(stars)
forest <- read_stars("https://www.massey.ac.nz/~jcmarsha/data/nz_map/forest.tif")
forest
```

## Forest cover

:::: columns
::: column
```{r forest1}
#| eval: false
ggplot(nz2023) +
  geom_sf(linewidth=0.1) +
  geom_stars(data=forest) +
  scale_fill_viridis_c(
    na.value='transparent'
    ) +
  labs(fill=NULL,
       title='Forest cover')
```

:::

::: {.column .absolute top="0" right="-5%"}
```{r}
#| echo: false
#| ref.label: forest1
```
:::
::::


## Forest cover

:::: columns
::: column
```{r forest2}
#| eval: false
lims <- nz2023 |>
  filter(sa3_name == "Stewart Island") |>
  st_bbox()

ggplot(nz2023) +
  geom_sf(linewidth=0.1) +
  geom_stars(data=forest) +
  scale_fill_viridis_c(
    na.value='transparent'
    ) +
  coord_sf(xlim=lims[c('xmin', 'xmax')],
           ylim=lims[c('ymin', 'ymax')]) +
  labs(fill=NULL,
       title="Forest cover")
```

:::

::: {.column .absolute top="0" right="-5%"}
```{r}
#| echo: false
#| ref.label: forest2
```
:::
::::

## Spatial joins

We can join datasets together with a spatial join:

```{r}
forest_sf <- forest |> st_as_sf() |> st_transform(2193)
sa3_forest <- nz2023 |> st_join(forest_sf)
sa3_forest |> select(sa3, sa3_name, forest.tif)
```

## Spatial join

We can then aggregate across the rows of the same geometry with the usual `group_by`/`summarise`:

```{r}
sa3_cover <- sa3_forest |>
  group_by(sa3, sa3_name, territorial_authority) |>
  summarise(forest = mean(forest.tif, na.rm=TRUE))
sa3_cover
```

## Forest cover

:::: columns
::: column
```{r forest3}
#| eval: false
sa3_cover |>
  ggplot() +
  aes(fill=forest) +
  geom_sf(linewidth=0.1) +
  scale_fill_viridis_c() +
  labs(fill=NULL,
       title="Forest cover")
```

:::

::: {.column .absolute top="0" right="-5%"}
```{r}
#| echo: false
#| ref.label: forest3
```
:::
::::


## Forest cover

:::: columns
::: column
```{r forest4}
#| eval: false
sa3_cover |>
  filter(territorial_authority == "Auckland") |>
  ggplot() +
  aes(fill=forest) +
  geom_sf(linewidth=0.1) +
  scale_fill_viridis_c() +
  labs(fill=NULL,
       title="Forest cover")
```

:::

::: {.column .absolute top="0" right="-5%"}
```{r}
#| echo: false
#| ref.label: forest4
```
:::
::::

## Further reading

- [Maps](https://ggplot2-book.org/maps.html) from ggplot2: Elegant Graphics for Data Analysis

- [Draw maps](http://vissoc.co/07-maps.html) from Data Visualisation A Practical Introduction

- [Introduction to sf and stars](https://r-spatial.org/book/07-Introsf.html) from Spatial Data Science With Applications in R

- Take 233.214 GIS and Spatial Statistics!
