---
title: "Amounts and proportions"
format:
  revealjs:
    width: 1280
    height: 720
    margin: 0.1
    echo: false
    message: false
    fig-width: 7
    fig-height: 7
---

## Amounts

```{r}
#| echo: false

library(tidyverse)
theme_set(theme_minimal(base_size=20))

ag_production <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/nz/ag_production.csv")
population <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/nz/nz_population.csv")
```

Often we have data that are as simple as some amounts.

:::{.fragment}

:::: {.columns}

::: {.column}
**Example:** Number of production animals in 2025.
:::

:::{.column}
```{r}
animals <- 
  ag_production |>
  slice_max(year) |> select(-year) |>
  mutate(value = value/1e6)

knitr::kable(animals, digits=1)
```
:::

::::

:::

## Bar charts are usually best

```{r}
#| fig.width: 10
#| fig.height: 6
#| fig.align: 'center'
ggplot(animals) +
  aes(x=fct_rev(fct_reorder(animal, value)), y=value) +
  geom_col(width=0.6, fill='steelblue') +
  scale_y_continuous(expand=expansion(mult = c(0,0.05))) +
  labs(x=NULL, y="Number of animals (m)") +
  theme(axis.ticks.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank())
```

## Long labels on angles?

```{r}
#| fig.width: 10
#| fig.height: 6
#| fig.align: 'center'
ggplot(animals) +
  aes(x=fct_rev(fct_reorder(animal, value)), y=value) +
  geom_col(width=0.6, fill='steelblue') +
  scale_y_continuous(expand=expansion(mult = c(0,0.05))) +
  labs(x=NULL, y="Number of animals (m)") +
  annotate('text', x=6.5, y=25, label="yeah nah", col='darkred', size=10) +
  theme(axis.ticks.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 30, hjust=1))
```

## We have the power of horizontal bars

```{r}
#| fig.width: 10
#| fig.height: 6
#| fig.align: 'center'
ggplot(animals) +
  aes(y=fct_reorder(animal, value), x=value) +
  geom_col(width=0.6, fill='steelblue') +
  scale_x_continuous(expand=expansion(mult = c(0,0.05))) +
  labs(y=NULL, x="Number of animals (m)") +
  annotate('text', x=20, y=1, label="yeah nah yeah", col='green4', size=10) +
  theme(axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank())
```

## Points are an option

```{r}
#| fig.width: 10
#| fig.height: 6
#| fig.align: 'center'
lego <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/lego/lego_sets.csv")
sets_per_year <- lego |>
  count(theme, year) |>
  summarise(sets=median(n), .by=theme) |>
  filter(sets > 5)
ggplot(sets_per_year) +
  aes(y=fct_reorder(theme, sets), x=sets) +
  geom_point(col='steelblue', size=5) +
  scale_x_continuous(limits=c(0, 51), expand=expansion(mult = c(0,0.05))) +
  labs(y=NULL, x="Median sets per year") +
#  annotate('text', x=20, y=1, label="yeah nah yeah", col='green4', size=10) +
  theme(axis.ticks.y = element_blank(),
        panel.grid.minor = element_blank())
```

## Bars should start at zero

```{r}
#| fig.width: 10
#| fig.height: 6
#| fig.align: 'center'
lego <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/lego/lego_sets.csv")
sets_per_year <- lego |>
  count(theme, year) |>
  summarise(sets=median(n), .by=theme) |>
  filter(sets > 5)
ggplot(sets_per_year) +
  aes(y=fct_reorder(theme, sets), x=sets) +
  geom_col(fill='steelblue', width=0.6) +
  scale_x_continuous(limits=c(0, 51), expand=expansion(mult = c(0,0.05))) +
  coord_cartesian(xlim=c(10, 51)) +
  labs(y=NULL, x="Median sets per year") +
  annotate('text', x=45, y=1, label="yeah nah", col='darkred', size=10) +
  theme(axis.ticks.y = element_blank(),
        panel.grid.minor = element_blank())
```

## Use points if zero is less important

```{r}
#| fig.width: 10
#| fig.height: 6
#| fig.align: 'center'
ggplot(sets_per_year) +
  aes(y=fct_reorder(theme, sets), x=sets) +
  geom_point(col='steelblue', size=5) +
  scale_x_continuous(limits=c(10, 51), expand=expansion(mult = c(0,0.05))) +
  labs(y=NULL, x="Median sets per year") +
  annotate('text', x=45, y=1, label="yeah nah yeah", col='green4', size=10) +
  theme(axis.ticks.y = element_blank(),
        panel.grid.minor = element_blank())
```

## Choose the 'right' order

:::: {.columns}

::: {.column}
```{r order1}
#| eval: false
#| echo: true
ggplot(sets_per_year) +
  aes(y=theme,
      x=sets) +
  geom_col()
```

The default order is alphabetical.

The `y` axis points up.
:::

:::{.column}
```{r}
#| ref.label: 'order1'
```
:::

::::

## Choose the 'right' order

:::: {.columns}

::: {.column}
```{r order2}
#| eval: false
#| echo: true
ggplot(sets_per_year) +
  aes(y=fct_reorder(theme, sets),
      x=sets) +
  geom_col()
```

`fct_reorder` will reorder the
grouping variable by the measure.
:::

:::{.column}
```{r}
#| ref.label: 'order2'
```
:::

::::


## Choose the 'right' order

:::: {.columns}

::: {.column}
```{r order3}
#| eval: false
#| echo: true
ggplot(sets_per_year) +
  aes(x=fct_reorder(theme, sets),
      y=sets) +
  geom_col()
```

Ascending order not as good for `x`
:::

:::{.column}
```{r}
#| ref.label: 'order3'
```
:::

::::

## Choose the 'right' order

:::: {.columns}

::: {.column}
```{r order4}
#| eval: false
#| echo: true
ggplot(sets_per_year) +
  aes(x=fct_rev(
          fct_reorder(theme, sets)
        ),
      y=sets) +
  geom_col()
```

`fct_rev` to reverse.

Alternatively `fct_reorder(.desc=TRUE)`
:::

:::{.column}
```{r}
#| ref.label: 'order4'
```
:::

::::

## Grouped bars

```{r}
set_counts <- lego |>
  count(theme, year, name = 'sets') |>
  filter(year > 2015, year < 2021) |>
  group_by(theme) |>
  filter(sum(sets) > 30) |>
  mutate(year = as_factor(year))
options('ggplot2.discrete.fill' = scales::viridis_pal(option="rocket")(6))
```

:::: {.columns}

::: {.column}
```{r group1}
#| eval: false
#| echo: true
ggplot(set_counts) +
  aes(x=theme,
      y=sets,
      fill=year) +
  geom_col()
```

Side by side often better
:::

:::{.column}
```{r}
#| ref.label: 'group1'
```
:::

::::

## Grouped bars


:::: {.columns}

::: {.column}
```{r group2}
#| eval: false
#| echo: true
ggplot(set_counts) +
  aes(x=theme,
      y=sets,
      fill=year) +
  geom_col(position='dodge')
```

Side by side often better
:::

:::{.column}
```{r}
#| ref.label: 'group2'
```
:::

::::


## Grouped bars

:::: {.columns}

::: {.column}
```{r group3}
#| eval: false
#| echo: true
ggplot(set_counts) +
  aes(x=year,
      y=sets,
      fill=theme) +
  geom_col(position='dodge')
```

Can swap `fill` and `x`
:::

:::{.column}
```{r}
#| ref.label: 'group3'
```
:::

::::

## Grouped bars

:::: {.columns}

::: {.column}
```{r group4}
#| eval: false
#| echo: true
ggplot(set_counts) +
  aes(x=year,
      y=sets) +
  geom_col() +
  facet_wrap(vars(theme))
```

Facetting is another option
:::

:::{.column}
```{r}
#| ref.label: 'group4'
```
:::

::::

## Counting first?

In the above charts we utilised `geom_col()` as our data were already summarised into counts.

::: {.fragment}
If your data aren't summarised, then you can:

::: {.incremental}
- Summarise it first with `dplyr::count()`.

- Use `geom_bar()` so that `stat_count()` does it for you.
:::
:::

## Counting first?

Consider the uncounted lego data for 2020:

```{r}
uncounted <- lego |> filter(year == 2020) |>
  add_count(theme) |> filter(n > 6) |> select(set, name, theme)
uncounted
```

## Counting first?

:::: {.columns}

::: {.column}
```{r count1}
#| eval: false
#| echo: true
ggplot(uncounted) +
  aes(y=theme) +
  geom_bar()
```

`geom_bar` uses `stat_count()` to count each `theme` to get `x`.
:::

:::{.column}
```{r}
#| ref.label: 'count1'
```
:::

::::

## Counting first?

:::: {.columns}

::: {.column}
```{r count2}
#| eval: false
#| echo: true
ggplot(uncounted) +
  aes(y=fct_infreq(theme)) +
  geom_bar()
```

`fct_infreq()` will order by frequency.
:::

:::{.column}
```{r}
#| ref.label: 'count2'
```
:::

::::

## Counting first?

:::: {.columns}

::: {.column}
```{r count2b}
#| eval: false
#| echo: true
ggplot(uncounted) +
  aes(y=fct_rev(
          fct_infreq(theme)
        )) +
  geom_bar()
```

Reversing is required for `y`.
:::

:::{.column}
```{r}
#| ref.label: 'count2b'
```
:::

::::

## Counting first?

:::: {.columns}

::: {.column}
```{r count3}
#| eval: false
#| echo: true
uncounted |>
  count(theme) |>
  ggplot() +
  aes(y=theme, x=n) +
  geom_col()
```

Alternatively we can count first and use `geom_col()`
:::

:::{.column}
```{r}
#| ref.label: 'count3'
```
:::

::::

## Counting first?

:::: {.columns}

::: {.column}
```{r count4}
#| eval: false
#| echo: true
uncounted |>
  count(theme) |>
  ggplot() +
  aes(y=fct_reorder(theme, n),
      x=n) +
  geom_col()
```

And use `fct_reorder()` for the count order.
:::

:::{.column}
```{r}
#| ref.label: 'count4'
```
:::

::::

# Proportions

## Proportions

The traditional chart for proportions of a whole is the pie.

::: {.fragment}
The problems with a pie are two fold:

::: {.incremental}
- Humans are only good at large differences or simple fractions (1/2, 1/4).

- "Hard" to do in `ggplot2`: use a bar with polar coordinates.
:::

:::

::: {.fragment}
Often a barchart (stacked or unstacked) is just as good. But not always!
:::

## Pie charts in ggplot2

```{r}
sets_2020 <- set_counts |> filter(year == 2020) |>
  ungroup() |>
  mutate(theme = fct_reorder(theme, sets))
```

:::: {.columns}

::: {.column}

```{r pie1}
#| eval: false
#| echo: true
ggplot(sets_2020) +
  aes(x=sets, y = "YY",
      fill=theme) +
  geom_col()
```

Start with a stacked bar chart.

Y axis is set to a single value so that `fill` stacks.
:::

:::{.column}
```{r}
#| ref.label: 'pie1'
```
:::

::::

## Pie charts in ggplot2

:::: {.columns}

::: {.column}

```{r pie2}
#| eval: false
#| echo: true
ggplot(sets_2020) +
  aes(x=sets, y = "YY",
      fill=theme) +
  geom_col() +
  coord_polar()
```

Switch to polar coordinates.
:::

:::{.column}
```{r}
#| ref.label: 'pie2'
```
:::

::::


## Pie charts in ggplot2

:::: {.columns}

::: {.column}

```{r pie3}
#| eval: false
#| echo: true
ggplot(sets_2020) +
  aes(x=sets, y = "YY",
      fill=theme) +
  geom_col() +
  coord_polar() +
  scale_x_continuous(name=NULL,
                   breaks=NULL) +
  scale_y_discrete(name=NULL,
                   breaks=NULL)
```

Drop the axis names and breaks.
:::

:::{.column}
```{r}
#| ref.label: 'pie3'
```
:::

::::


## Pie charts in ggplot2

:::: {.columns}

::: {.column}

```{r pie4}
#| eval: false
#| echo: true
ggplot(sets_2020) +
  aes(x=sets, y = "YY",
      fill=theme) +
  geom_col() +
  coord_polar() +
  scale_x_continuous(name=NULL,
                   breaks=NULL) +
  scale_y_discrete(name=NULL,
                   breaks=NULL) +
  theme(panel.grid=element_blank())
```

Remove panel grid.
:::

:::{.column}
```{r}
#| ref.label: 'pie4'
```
:::

::::


## Pie vs bar

```{r}
#| echo: false
#| fig.width: 12
#| fig.height: 7
#| fig.align: 'center'
library(patchwork)
pie <- ggplot(sets_2020) +
  aes(x=sets, y = "YY",
      fill=theme) +
  geom_col() +
  coord_polar() +
  scale_x_continuous(name=NULL,
                   breaks=NULL) +
  scale_y_discrete(name=NULL,
                   breaks=NULL) +
  theme(panel.grid=element_blank())

bar <- ggplot(sets_2020) +
  aes(x=sets, y=theme, fill=theme) +
  geom_col() +
  scale_x_continuous(expand=expansion(mult=c(0,0.05))) +
  guides(fill='none') +
  theme(panel.grid.major.y=element_blank(),
        panel.grid.minor=element_blank()) +
  labs(y=NULL, x = "Number of sets")

pie + bar
```

## Making bars prettier

:::: {.columns}

::: {.column}

```{r pretty1}
#| eval: false
#| echo: true
ggplot(sets_2020) +
  aes(x=sets, y=theme) +
  geom_col()
```

Start simple.
:::

:::{.column}
```{r}
#| ref.label: 'pretty1'
```
:::

::::

## Making bars prettier

:::: {.columns}

::: {.column}

```{r pretty2}
#| eval: false
#| echo: true
ggplot(sets_2020) +
  aes(x=sets, y=theme) +
  geom_col() +
  scale_x_continuous(
    expand=expansion(mult=c(0,0.05))
    )
```

Change `expand` so the axis is at 0.
:::

:::{.column}
```{r}
#| ref.label: 'pretty2'
```
:::

::::

## Making bars prettier

:::: {.columns}

::: {.column}

```{r pretty3}
#| eval: false
#| echo: true
ggplot(sets_2020) +
  aes(x=sets, y=theme, fill=theme) +
  geom_col() +
  scale_x_continuous(
    expand=expansion(mult=c(0,0.05))
    ) +
  guides(fill='none')
```

Add some colour.

Turn off the guide.
:::

:::{.column}
```{r}
#| ref.label: 'pretty3'
```
:::

::::


## Making bars prettier

:::: {.columns}

::: {.column}

```{r pretty4}
#| eval: false
#| echo: true
ggplot(sets_2020) +
  aes(x=sets, y=theme, fill=theme) +
  geom_col() +
  scale_x_continuous(
    expand=expansion(mult=c(0,0.05))
    ) +
  guides(fill='none') +
  labs(y=NULL, x = "Number of sets")
```

Improve the labels.
:::

:::{.column}
```{r}
#| ref.label: 'pretty4'
```
:::

::::

## Making bars prettier

:::: {.columns}

::: {.column}

```{r pretty5}
#| eval: false
#| echo: true
ggplot(sets_2020) +
  aes(x=sets, y=theme, fill=theme) +
  geom_col() +
  scale_x_continuous(
    expand=expansion(mult=c(0,0.05))
    ) +
  guides(fill='none') +
  labs(y=NULL, x = "Number of sets") +
  theme(panel.grid.major.y=element_blank(),
        panel.grid.minor=element_blank())
```

Tweak the theme to remove some of the grid.
:::

:::{.column}
```{r}
#| ref.label: 'pretty5'
```
:::

::::


## Making bars prettier

:::: {.columns}

::: {.column}

```{r pretty6}
#| eval: false
#| echo: true
ggplot(sets_2020) +
  aes(x=sets, y=theme, fill=theme) +
  geom_col(width=0.8) +
  scale_x_continuous(
    expand=expansion(mult=c(0,0.05))
    ) +
  guides(fill='none') +
  labs(y=NULL, x = "Number of sets") +
  theme(panel.grid.major.y=element_blank(),
        panel.grid.minor=element_blank())
```

Reduce the bar width.
:::

:::{.column}
```{r}
#| ref.label: 'pretty6'
```
:::

::::
