---
title: "Colours"
format:
  revealjs:
    theme: [solarized, slides.scss]
    width: 1280
    height: 720
    margin: 0.1
    echo: false
    message: false
    fig-width: 7
    fig-height: 7
---

```{r}
#| echo: false

library(tidyverse)
theme_set(theme_minimal(base_size=20))

ag_production <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/nz/ag_production.csv")
population <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/nz/nz_population.csv")
```

# Colour spaces

## RGB colour space

The default colour space for most computer things is RGB:

- red R: amount of red, usually 0-255.
- red G: amount of green, usually 0-255.
- red B: amount of blue, usually 0-255.

:::{.fragment}
This is a strange scale that is hard to reason about.
:::

## RGB colour space

:::{.incremental}
- The 0-255 scale on each of the components is due to computers holding value in memory in bytes.

- A byte is 8 binary bits of 0/1 so can represent $2^8=256$ different combinations (0 through 255).

- Typically it is represented in **hexadecimal** or base 16, where the digits are 0-9 and A-F (for 10 through 15).

- e.g. #FF007F would represent R=255, G=0, B=127.
:::

## RGB colour space

```{r}
#| fig.width: 10
#| fig.align: 'center'
expand_grid(red=seq(0,256,by=32), green=seq(0,256,by=2), blue=seq(0,256,by=2)) |>
  mutate(col=rgb(red,green,blue,maxColorValue=256)) |>
  ggplot() +
  aes(x=green, y=blue, fill=col) +
  geom_tile() +
  scale_fill_identity() +
  facet_wrap(vars(red), labeller=label_both) +
  theme(panel.grid.minor=element_blank())
```

## HSV colour space

Colours are described by their

- Hue H (0-360): the hue (from a colour wheel).

- Saturation S (0-1): the intensity of the colour for a given brightness.

- Value V (0-1): how much light is emitted.

:::{.fragment}
Unlike RGB which is a cartesian colour space, HSV is a cylindrical space (polar coordinates for hue and saturation)
:::

## HSV colour space

```{r}
#| fig.width: 10
#| fig.align: 'center'
library(colorspace)
expand_grid(hue=seq(0,320,by=40)+20, saturation=seq(0,1,length.out=100), value=seq(0,1,length.out=100)) |>
  mutate(col=HSV(hue, saturation, value) |> hex()) |>
  ggplot() +
  aes(x=value, y=saturation, fill=col) +
  geom_tile() +
  scale_fill_identity() +
  facet_wrap(vars(hue), labeller=label_both) +
  theme(panel.grid.minor=element_blank())
```

## HLS colour space

HLS is related to HSV: instead of 'value' it has 'lightness':

- Hue H (0-360): the hue (from a colour wheel).

- Lightness L (0-1): how bright it is compared to white.

- Saturation S (0-1): the intensity of the colour for a given brightness.

## HLS colour space

```{r}
#| fig.width: 10
#| fig.align: 'center'
library(colorspace)
expand_grid(hue=seq(0,320,by=40)+20, lightness=seq(0,1,length.out=100), saturation=seq(0,1,length.out=100)) |>
  mutate(col=HLS(hue, lightness, saturation) |> hex()) |>
  ggplot() +
  aes(x=lightness, y=saturation, fill=col) +
  geom_tile() +
  scale_fill_identity() +
  facet_wrap(vars(hue), labeller=label_both) +
  theme(panel.grid.minor=element_blank())
```

:::{.fragment}
Encodes hue directly but HSV or HLS luminance doesn't match how we see colour.
:::

## LUV colour space

LUV uses a non-linear encoding of colour in an **opponent** space:

- Luminance L (0-100): amount of light emitted

- Green/red balance u (-100 to 100): amount of green vs red

- Blue/yellow balance v (-100 to 100): amount of blue vs yellow

:::{.fragment}
LUV's luminance operates very similarly to human perception of luminance.
:::

## LUV colour space

```{r}
#| fig.width: 10
#| fig.align: 'center'
expand_grid(luminance=seq(10,90,length.out=9), u=seq(-120,120,length.out=100), v=seq(-120,120,length.out=100)) |>
  mutate(col=LUV(luminance, u, v) |> hex()) |>
  ggplot() +
  aes(x=u, y=v, fill=col) +
  geom_tile() +
  scale_fill_identity() +
  facet_wrap(vars(luminance), labeller=label_both) +
  labs(x = 'green/red U', y='blue/yellow V') +
  theme(panel.grid.minor=element_blank())
```

:::{.fragment}
The space is confined to what most devices can represent.
:::

## HCL color space

The disadvantage of LUV is it doesn't have a single parameter for hue (colour).

:::{.fragment}
HCL solves this:

- Hue H (0-360): the hue from a colour wheel.

- Chroma C (approx 0-150): the colour intensity.

- Luminance L (0-100): how bright the colour is compared with white.
:::

## HCL color space

```{r}
#| fig.width: 10
#| fig.align: 'center'
expand_grid(hue=seq(0,320,length.out=9)+20, chroma=seq(0, 150), luminance=1:100) |>
  mutate(col=hcl(hue, chroma, luminance, fixup=FALSE)) |>
  ggplot() +
  aes(x=chroma, y=luminance, fill=col) +
  geom_tile() +
  scale_fill_identity() +
  facet_wrap(vars(hue), labeller=label_both) +
  theme(panel.grid.minor=element_blank())
```

:::{.fragment}
This space is usually the most intuitive for humans to reason about.
:::

##

<div>
```{=html}
<iframe src="https://shiny.massey.ac.nz/jcmarsha/hclcolorpicker" width=960 height=720></iframe>
```
</div>

[`color_picker()`<br>from `colorspace`]{.absolute right=0 top=200}

# Colour blindness

## Colour blindness is a thing

```{r}
#| fig.width: 9
#| fig.align: 'center'
cols = darken(scales::brewer_pal(palette='RdYlGn')(5), amount=0.04)
swatchplot(cols, cvd=TRUE, border='transparent', cex=2)
```

Dueteranomaly (red-green colour blind) is the most common.

## Choose colours with this in mind

```{r}
#| fig.width: 9
#| fig.align: 'center'
cols = darken(scales::brewer_pal(palette='PiYG')(5), amount=0.04)
swatchplot(cols, cvd=TRUE, border='transparent', cex=2)
```

Use `colorspace::swatchplot(cols, cvd=TRUE)` to simulate.

# Colour scales

## Purpose of colour

:::{.incremental}
- Distinguish categories (qualitative scales)

- Represent numeric value (sequential)

- Represent numeric value (diverging)

- Highlight a set of observations.
:::

## Distinguishing categories

```{r}
animals <- ag_production |> filter(year > 1995) |>
  mutate(value = value/1e6) |> filter(value > 3)
```

::::{.columns}

:::{.column}
```{r qual1}
#| echo: true
#| eval: false
ggplot(animals) +
  aes(x=year, y=value, col=animal)+
  geom_line(linewidth=2) +
  scale_color_brewer(palette="Set1")
```

"Set1" from Color Brewer
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'qual1'
```
:::

::::

## Distinguishing categories

::::{.columns}

:::{.column}
```{r qual2}
#| echo: true
#| eval: false
ggplot(animals) +
  aes(x=year, y=value, col=animal)+
  geom_line(linewidth=2) +
  scale_color_brewer(palette="Dark2")
```

"Dark2" from Color Brewer
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'qual2'
```
:::

::::


## Distinguishing categories

::::{.columns}

:::{.column}
```{r qual3}
#| echo: true
#| eval: false
ggplot(animals) +
  aes(x=year, y=value, col=animal)+
  geom_line(linewidth=2) +
  scale_color_manual(
    values=c(Sheep="#E69F00",
             `Beef Cattle`="#0072B2",
             `Dairy Cattle`="#009E73",
             Poultry="#F0E442")
  )
```

Picking some colours from the Okabe-Ito palette.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'qual3'
```
:::

::::

## Sequential values

```{r}
births <- read_csv(here::here("data/births_tidy.csv"))
```

::::{.columns}

:::{.column}
```{r seq1}
#| echo: true
#| eval: false
ggplot(births) +
  aes(x=year, y=age, fill=births) +
  geom_tile()
```

Default is blue to black.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'seq1'
```
:::

::::

## Sequential values

::::{.columns}

:::{.column}
```{r seq2}
#| echo: true
#| eval: false
ggplot(births) +
  aes(x=year, y=age, fill=births) +
  geom_tile() +
  scale_fill_gradient(low='grey30', high='orange')
```

You can specify the end points of a gradient.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'seq2'
```
:::

::::

## Sequential values

::::{.columns}

:::{.column}
```{r seq3}
#| echo: true
#| eval: false
ggplot(births) +
  aes(x=year, y=age, fill=births) +
  geom_tile() +
  scale_fill_binned(low='grey30', high='orange',
                    n.breaks=5)
```

Binning the guide can be useful.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'seq3'
```
:::

::::


## Sequential values

::::{.columns}

:::{.column}
```{r seq4}
#| echo: true
#| eval: false
ggplot(births) +
  aes(x=year, y=age, fill=births) +
  geom_tile() +
  scale_fill_viridis_c()
```

Viridis is colorblind safe.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'seq4'
```
:::

::::

## Sequential values

::::{.columns}

:::{.column}
```{r seq5}
#| echo: true
#| eval: false
ggplot(births) +
  aes(x=year, y=age, fill=births) +
  geom_tile() +
  scale_fill_viridis_b()
```

`viridis_b` for a binned version.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'seq5'
```
:::

::::


## Sequential values

::::{.columns}

:::{.column}
```{r seq6}
#| echo: true
#| eval: false
ggplot(births) +
  aes(x=year, y=age, fill=births) +
  geom_tile() +
  scale_fill_viridis_c(option="mako")
```

There are 8 different options.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'seq6'
```
:::

::::

## Diverging values

```{r}
solar <- read_csv(here::here("data/solar_hourly_surplus.csv")) |>
  mutate(surplus = (produced-consumed)/1000)
```

::::{.columns}

:::{.column}
```{r div1}
#| echo: true
#| eval: false
ggplot(solar) +
  aes(x=day, y=hour, fill=surplus) +
  geom_tile() +
  scale_fill_distiller(palette="RdYlBu")
```

`scale_fill_distiller` extends the ColorBrewer scales to continuous.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'div1'
```
:::

::::


## Diverging values

::::{.columns}

:::{.column}
```{r div2}
#| echo: true
#| eval: false
ggplot(solar) +
  aes(x=day, y=hour, fill=surplus) +
  geom_tile() +
  scale_fill_fermenter(
    palette="RdYlBu",
    n.breaks=5
    )
```

`scale_fill_fermenter` bins colours.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'div2'
```
:::

::::

## Highlighting

::::{.columns}

:::{.column}
```{r high1}
#| echo: true
#| eval: false
animals |>
  mutate(highlight=fct_collapse(
    animal, 
    Sheep='Sheep',
    other_level="Others")) |>
  ggplot() +
  aes(x=year, y=value,
      group=animal,
      col=highlight) +
  geom_line(linewidth=2) +
  scale_colour_manual(
    values=c(Sheep = "darkorange",
             Others="grey30")
  )
```

Highlighting is best done with a new column.
:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label: 'high1'
```
:::

::::
