---
title: "Distributions"
format:
  revealjs:
    theme: [solarized, slides.scss]
    width: 1280
    height: 720
    margin: 0.1
    smaller: false
    echo: false
    message: false
    fig-width: 7
    fig-height: 7
---

```{r}
#| echo: false

library(tidyverse)
theme_set(theme_minimal(base_size=20))

ag_production <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/nz/ag_production.csv")
population <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/nz/nz_population.csv")
```

## Distributions

There are a number of options for visualising distributions:

:::{.incremental}
- Histograms
- Densities
- Boxplots
- Violins (densities again)
- Dots (jittered scatter, dot plot)
:::

:::{.fragment}
Except for the jittered scatter, these display a **statistical summary** of the data.
:::

## Petrels caught as by-catch in NZ waters{.smaller}

```{r}
petrels <- read_csv("https://www.massey.ac.nz/~jcmarsha/227212/data/petrels.csv") |> select(year=Year, month=Month, area=Area, sex=Sex, wing_length=R.Wing.Lth, tarsus_length=R.Tarsus.Lth) |>
  na.omit()
petrels |>
  slice_head(n=10) |>
  knitr::kable()
```
## Histograms: Bin and count{.smaller}

::::{.columns}

```{r}
petrel_counts <- petrels |> mutate(range = cut(wing_length, breaks=seq(340,420,by=5), labels=FALSE)) |>
  mutate(`wing range` = paste(seq(340,415,by=5)[range], seq(345,420,by=5)[range], sep="-")) |>
  count(`wing range`, range, name='count') |>
  arrange(`wing range`)
```

:::{.column width="25%"}
```{r}
petrel_counts |> filter(range <= 9) |> select(-range) |> knitr::kable()
```
:::

:::{.column width="25%"}
```{r}
petrel_counts |> filter(range > 9) |> select(-range) |> knitr::kable()
```
:::

:::{.column width="50%"}

:::{.fragment}

```{r}
#| fig.height: 5
ggplot(petrels) +
  aes(x=wing_length) +
  geom_histogram(breaks = seq(340,420,by=5),colour='white', fill='darkorange') +
  scale_y_continuous(expand=expansion(mult=c(0,0.05))) +
  labs(x="Wing length (mm)")
```
:::

:::

::::

## Depends on chosen bins

::::{.columns}

:::{.column}
```{r bin1}
#| eval: false
#| echo: true
ggplot(petrels) +
  aes(x=wing_length) +
  geom_histogram(bins=50)
```
:::

:::{.column}
```{r}
#| fig.height: 5
#| ref.label: "bin1"
```
:::

::::

## Depends on chosen bins

::::{.columns}

:::{.column}
```{r bin2}
#| eval: false
#| echo: true
ggplot(petrels) +
  aes(x=wing_length) +
  geom_histogram(bins=25)
```
:::

:::{.column}
```{r}
#| fig.height: 5
#| ref.label: "bin2"
```
:::

::::


## Depends on chosen bins

::::{.columns}

:::{.column}
```{r bin3}
#| eval: false
#| echo: true
ggplot(petrels) +
  aes(x=wing_length) +
  geom_histogram(bins=5)
```
:::

:::{.column}
```{r}
#| fig.height: 5
#| ref.label: "bin3"
```
:::

::::


## Depends on chosen bins

::::{.columns}

:::{.column}
```{r bin4}
#| eval: false
#| echo: true
ggplot(petrels) +
  aes(x=wing_length) +
  geom_histogram(bins=16)
```
:::

:::{.column}
```{r}
#| fig.height: 5
#| ref.label: "bin4"
```
:::

::::


## Depends on chosen bins

::::{.columns}

:::{.column}
```{r bin5}
#| eval: false
#| echo: true
ggplot(petrels) +
  aes(x=wing_length) +
  geom_histogram(binwidth=5)
```

We can specify `binwidth` instead.
:::

:::{.column}
```{r}
#| fig.height: 5
#| ref.label: "bin5"
```
:::

::::


## Depends on chosen bins

::::{.columns}

:::{.column}
```{r bin6}
#| eval: false
#| echo: true
ggplot(petrels) +
  aes(x=wing_length) +
  geom_histogram(binwidth=5,
                 boundary=340)
```

We can specify `binwidth` instead.

And position exactly with `boundary` or `center`.
:::

:::{.column}
```{r}
#| fig.height: 5
#| ref.label: "bin6"
```
:::

::::

## Densities

Kernel density estimates are a smooth alternative.

:::{.fragment}
A kernel density estimates the **probability density** by averaging the contribution of kernels centered on each data point.
:::

:::{.fragment}
$$
f(x) = \frac{1}{n}\sum_{i=1}^n K_h(x - x_i)
$$
:::

:::{.fragment}
The kernel $K_h$ is often a normal distribution with standard deviation $h$.
:::

:::{.fragment}
The **bandwidth** $h$ determines the smoothness of the estimate
:::

## Densities

:::: {.columns}

::: {.column}
```{r}
#| fig.height: 5
ggplot(petrels) +
  aes(x=wing_length) +
  geom_histogram(breaks = seq(340,420,by=5),colour='white', fill='darkorange') +
  scale_y_continuous(expand=expansion(mult=c(0,0.05))) +
  labs(x="Wing length (mm)")
```
:::

::: {.column}
::: {.fragment}
```{r}
#| fig.height: 5
ggplot(petrels) +
  aes(x=wing_length) +
  geom_density(colour='white', fill='darkorange') +
  scale_y_continuous(expand=expansion(mult=c(0,0.05))) +
  labs(x="Wing length (mm)")
```
:::

:::{.fragment}
The scale is proportions (area is 1)
:::

:::

::::

## Choosing bandwidth

:::: {.columns}

::: {.column}
```{r bw1}
#| eval: false
#| echo: true
ggplot(petrels) +
  aes(x=wing_length) +
  geom_density(fill='pink',
               adjust=0.5)
```

`adjust < 1` for less smoothing.
:::

::: {.column}
```{r}
#| fig.height: 5
#| ref.label: "bw1"
```
:::

::::

## Choosing bandwidth

:::: {.columns}

::: {.column}
```{r bw2}
#| eval: false
#| echo: true
ggplot(petrels) +
  aes(x=wing_length) +
  geom_density(fill='pink',
               adjust=2)
```

`adjust > 1` for more smoothing.
:::

::: {.column}
```{r}
#| fig.height: 5
#| ref.label: "bw2"
```
:::

::::

## Choosing bandwidth

:::: {.columns}

::: {.column}
```{r bw3}
#| eval: false
#| echo: true
ggplot(petrels) +
  aes(x=wing_length) +
  geom_density(fill='pink')
```

The default is usually pretty good.
:::

::: {.column}
```{r}
#| fig.height: 5
#| ref.label: "bw3"
```
:::

::::

## What about groups?

:::: {.columns}

::: {.column}
```{r groups1}
#| eval: false
#| echo: true
ggplot(petrels) +
  aes(x=wing_length, fill=sex) +
  geom_histogram(binwidth=5)
```

By default bars stack.

Not what we want!
:::

::: {.column}
```{r}
#| fig.height: 5
#| ref.label: "groups1"
```
:::

::::

## What about groups?

:::: {.columns}

::: {.column}
```{r groups2}
#| eval: false
#| echo: true
ggplot(petrels) +
  aes(x=wing_length, fill=sex) +
  geom_histogram(binwidth=5,
                 position='identity')
```

Use 'identity' to position all bars at zero.

One will be in front of the other.
:::

::: {.column}
```{r}
#| fig.height: 5
#| ref.label: "groups2"
```
:::

::::

## What about groups?

:::: {.columns}

::: {.column}
```{r groups3}
#| eval: false
#| echo: true
ggplot(petrels) +
  aes(x=wing_length, fill=sex) +
  geom_histogram(binwidth=5,
                 position='identity',
                 alpha=0.4)
```

Use 'identity' to position all bars at zero.

Possibly use transparency?
:::

::: {.column}
```{r}
#| fig.height: 5
#| ref.label: "groups3"
```
:::

::::

## What about groups?

:::: {.columns}

::: {.column}
```{r groups4}
#| eval: false
#| echo: true
ggplot(petrels) +
  aes(x=wing_length, fill=fct_rev(sex)) +
  geom_histogram(binwidth=5,
                 position='identity',
                 alpha=0.8)
```

Use 'identity' to position all bars at zero.

Reorder the groups?
:::

::: {.column}
```{r}
#| fig.height: 5
#| ref.label: "groups4"
```
:::

::::


## What about groups?

:::: {.columns}

::: {.column}
```{r groups5}
#| eval: false
#| echo: true
ggplot(petrels) +
  aes(x=wing_length) +
  geom_histogram(binwidth=5) +
  facet_wrap(vars(sex))
```

Using facets is probably best.
:::

::: {.column}
```{r}
#| fig.height: 5
#| ref.label: "groups5"
```
:::

::::

## Grouped densities are better

:::: {.columns}

::: {.column}
```{r groups6}
#| eval: false
#| echo: true
ggplot(petrels) +
  aes(x=wing_length, fill=sex) +
  geom_density()
```

Densities don't stack.
:::

::: {.column}
```{r}
#| fig.height: 5
#| ref.label: "groups6"
```
:::

::::

## Grouped densities are better

:::: {.columns}

::: {.column}
```{r groups7}
#| eval: false
#| echo: true
ggplot(petrels) +
  aes(x=wing_length, fill=sex) +
  geom_density(position='stack')
```

Stacking is not advised!
:::

::: {.column}
```{r}
#| fig.height: 5
#| ref.label: "groups7"
```
:::

::::

## Grouped densities are better

:::: {.columns}

::: {.column}
```{r groups8}
#| eval: false
#| echo: true
ggplot(petrels) +
  aes(x=wing_length, fill=sex) +
  geom_density(alpha=0.6)
```

A bit of transparency will work.

Densities are better than
:::

::: {.column}
```{r}
#| fig.height: 5
#| ref.label: "groups8"
```
:::

::::

## Many groups don't work so well

```{r}
#| fig.width: 10
#| fig.align: 'center'
ggplot(petrels) +
  aes(x=wing_length, fill=area) +
  geom_density(colour='white', alpha=0.5) +
  scale_y_continuous(expand=expansion(mult=c(0,0.05))) +
  labs(x="Wing length (mm)") +
  scale_fill_viridis_d()
```

## Facets work better

```{r}
#| fig.width: 10
#| fig.align: 'center'
ggplot(petrels) +
  aes(x=wing_length) +
  geom_density(colour='white', fill='darkorange') +
  scale_y_continuous(expand=expansion(mult=c(0,0.05))) +
  labs(x="Wing length (mm)") +
  facet_wrap(vars(area))
```

## Boxplots instead?

```{r}
#| fig.width: 10
#| fig.align: 'center'
ggplot(petrels) +
  aes(x=area, y=wing_length) +
  geom_boxplot(fill='orange', alpha=0.5) +
  labs(x="Area", y="Wing length (mm)")
```

:::{.fragment}
Side by side boxplots are excellent for many groups.
:::

## Violins?

```{r}
#| fig.width: 10
#| fig.align: 'center'
ggplot(petrels) +
  aes(x=area, y=wing_length) +
  geom_violin(fill='orange', alpha=0.5) +
  labs(x="Area", y="Wing length (mm)")
```

:::{.fragment}
Violins are back to back densities per group.
:::

## Scatterplots?

```{r}
#| fig.width: 10
#| fig.align: 'center'
ggplot(petrels) +
  aes(x=area, y=wing_length) +
  geom_point(position=position_jitter(height=0, width=0.3), alpha=0.6) +
  labs(x="Area", y="Wing length (mm)")
```

:::{.fragment}
Scatterplots with some horizontal jitter can work too.
:::

## Ridge lines

```{r}
#| fig.width: 10
#| fig.align: 'center'
library(ggdist)
ggplot(petrels) +
  aes(x=wing_length, y=area) +
  geom_slab(stat='slab', height=2, fill='darkorange', colour='white', trim=FALSE, density='unbounded') +
  scale_y_discrete(expand=expansion(mult=c(0,0.05))) +
  labs(x="Wing length (mm)", y="Area") +
  theme(axis.text.y = element_text(vjust=0))
```

:::{.fragment}
Offsetting densities on the y axis with `geom_slab()` from `ggdist`.
:::