---
title: "Design"
knitr:
  opts_chunk: 
    dev: "ragg_png"
format:
  revealjs:
    theme: [solarized, slides.scss]
    width: 1280
    height: 720
    margin: 0.1
    echo: false
    message: false
    fig-width: 7
    fig-height: 8
---

```{r}
#| echo: false

library(tidyverse)
library(patchwork)
theme_set(theme_minimal(base_size=20))
```

# Chart layout

## Chart layout

- We've already seen that facetting allows multiple charts of the same type to be composed into one image.

- What if we have disparate charts that we wish to display at once?

- The `patchwork` package is useful for this.

## Using patchwork

The `patchwork` package overrides arithmetic operators[^1] to combine `ggplot2` chart objects.

:::{.fragment}
```{r}
#| eval: false
#| echo: true
library(patchwork)

# code for plot 1 saved
plot1 <- ggplot(data=foo) + ...

# code for plot 2 saved
plot2 <- ggplot(data=bar) + ...

# code to combine
plot1 + plot2 # side-by-side
```
:::

[^1]: similar to how `ggplot2` overrides `+`

## Side bar: print is magic

Saving the result of `ggplot2` code not print the chart.

:::{.incremental}
- The assignment `plot1 <- ggplot(data=foo) + ...` executes the `ggplot2` expression and assigns the result to the `plot1` object.

- But this does not draw the chart!

- Instead it produces an object that tells R how to produce a chart when we need to do so.

- Why do we normally get a chart when we don't have assignment then?

- **Because the default R action when running an expression without assignment is to print the result.**

- The **print** method of `ggplot2` is what does the actual drawing.
:::

## Side by side

```{r patch1}
#| echo: true
#| fig.width: 10
#| fig.height: 5
#| fig.align: 'center'
g1 <- ggplot(penguins) + aes(x=flipper_len, y=body_mass, col=species) + geom_point()
g2 <- ggplot(penguins) + aes(x=species, fill=species) + geom_bar()
g1 + g2
```

## Above and below

```{r patch2}
#| echo: true
#| fig-width: 10
#| fig.height: 5
#| fig.align: 'center'
g1 <- ggplot(penguins) + aes(x=flipper_len, y=body_mass, col=species) + geom_point()
g2 <- ggplot(penguins) + aes(x=species, fill=species) + geom_bar()
g1 / g2
```

## Combinations

```{r patch3}
#| echo: true
#| fig-width: 10
#| fig.height: 5
#| fig.align: 'center'
g1 <- ggplot(penguins) + aes(x=flipper_len, y=body_mass, col=species) + geom_point()
g2 <- ggplot(penguins) + aes(x=species, fill=species) + geom_bar()
g3 <- ggplot(penguins) + aes(x=bill_len, fill=species) + geom_density(alpha=0.7)
g1 / (g2+g3)
```

## Gathering guides

```{r patch4}
#| echo: true
#| fig-width: 10
#| fig.height: 5
#| fig.align: 'center'
g1 <- ggplot(penguins) + aes(x=flipper_len, y=body_mass, col=species) + geom_point()
g2 <- ggplot(penguins) + aes(x=species, fill=species) + geom_bar()
g3 <- ggplot(penguins) + aes(x=bill_len, fill=species) + geom_density(alpha=0.7)
g1 / (g2+g3) + plot_layout(guides = 'collect')
```

## Gathering guides

```{r patch5}
#| echo: true
#| fig-width: 10
#| fig.height: 5
#| fig.align: 'center'
g1 <- ggplot(penguins) + aes(x=flipper_len, y=body_mass, col=species) + geom_point()
g2 <- ggplot(penguins) + aes(x=species, fill=species) + geom_bar()
g4 <- ggplot(penguins) + aes(x=bill_len, y=bill_dep, col=species) + geom_point()
g1 / (g2+g4) + plot_layout(guides = 'collect')
```

## Tuning width and height

```{r patch6}
#| echo: true
#| fig-width: 10
#| fig.height: 5
#| fig.align: 'center'
g1 <- ggplot(penguins) + aes(x=flipper_len, y=body_mass, col=species) + geom_point()
g2 <- ggplot(penguins) + aes(x=species, fill=species) + geom_bar()
g4 <- ggplot(penguins) + aes(x=bill_len, y=bill_dep, col=species) + geom_point()
g1 / (g2+g4 + plot_layout(widths=c(1, 3))) + plot_layout(guides = 'collect')
```

## Tuning width and height

```{r patch7}
#| echo: true
#| fig-width: 10
#| fig.height: 5
#| fig.align: 'center'
g1 <- ggplot(penguins) + aes(x=flipper_len, y=body_mass, col=species) + geom_point()
g2 <- ggplot(penguins) + aes(x=species, fill=species) + geom_bar()
g4 <- ggplot(penguins) + aes(x=bill_len, y=bill_dep, col=species) + geom_point()
g1 / (g2+g4 + plot_layout(widths=c(1, 3))) + plot_layout(heights=c(5,3), guides = 'collect')
```

## Theme modification

```{r patch8}
#| echo: true
#| fig-width: 10
#| fig.height: 5
#| fig.align: 'center'
g1 / (g2+g4) + plot_layout(widths=c(1, 3)) &
  theme_bw(base_size=20)
```

# Themes

## Theming

:::{.incremental}
- Themes in `ggplot2` are governed by the `theme_*()` functions.

- Individual elements can be modified using `theme()`.

- Everything you see can be modified.

- Good figure design often requires a bunch of customisation.
:::

## Theming

```{r}
#| echo: true
#| fig-width: 10
#| fig.height: 5
#| fig.align: 'center'
g1 + theme_grey()
```


## Theming

```{r}
#| echo: true
#| fig-width: 10
#| fig.height: 5
#| fig.align: 'center'
g1 + theme_bw()
```


## Theming

```{r}
#| echo: true
#| fig-width: 10
#| fig.height: 5
#| fig.align: 'center'
g1 + theme_dark()
```

## Theming

```{r}
#| echo: true
#| fig-width: 10
#| fig.height: 5
#| fig.align: 'center'
g1 + theme_minimal()
```


## Theming: ggthemes

```{r}
#| echo: true
#| fig-width: 10
#| fig.height: 5
#| fig.align: 'center'
library(ggthemes)
g1 + theme_economist()
```

## Theming: ggthemes (NOPE!)

```{r}
#| echo: true
#| fig-width: 10
#| fig.height: 5
#| fig.align: 'center'
library(ggthemes)
g1 + theme_excel() + scale_colour_excel()
```

# Fonts

## Fonts and typefaces

The 'language' of fonts is a bit confusing!

:::{.incremental}
- A **typeface** is a style of lettering that forms a cohesive whole.

- Typically it combines different weights (bold, normal, light) as well as stylistic changes (italics, upright, small caps).

- A **font** is a subset of a typeface that describes a particular variation. It's a combination of width, weight and style/slant.

- This distinction doesn't really matter too much.
:::

## Digital font files

Fonts are stored on your computer in a number of different formats:

:::{.incremental}

- TrueType (ttf/ttc). Developed by Apple in the 1980s, it's what all modern fonts are based on.

- OpenType (otf/otc). Developed by Microsoft and Adobe to improve upon TrueType. Introduced more glyphs (letters) and additional features for **shaping**.

- Web Open Font Format (woff/woff2). A compression wrapper around TrueType/OpenType and reduces the number of opentype features to ensure download sizes are minimal for web browsing. Limited support outside of browsers.
:::

## Digital font files

Where fonts are stored depends on your system. The `systemfonts` package is designed to find them and make them usable from R:

```{r}
#| echo: true
systemfonts::system_fonts() |> select(name, family, style, weight, path)
```

## Limitations of fonts in ggplot2

:::{.incremental}
- In `ggplot2` the type face is known as `family` and the font is `face` (or `fontface`).

- The main limitation is that `face` or `fontface` can take only four options: plain, bold, italic and bold-italic.

- To get an 'extra bold' or 'light' version of a typeface we need to load it as a separate `family`.
:::

## Limitations of fonts in the graphics device

:::{.incremental}
- The **graphics device** is the library and code that draws stuff on a screen, or into a file.

- The choice of font file to use, how the text should be 'shaped' (spacing, ligatures etc) and sized is up to the graphics device.

- The graphics device on a Mac is different to the one on Windows!

- There are at least 3 different graphics devices available for drawing a plot to a .png file!
:::

## Best graphic devices to use for nice fonts

:::{.incremental}
- Use the `ragg` device along with for rendering raster (i.e. pixels) graphics.

- Use the `svglite` device for rendering vector graphics (SVG) for the web.

- In RStudio, set "AGG" as the backend under Global Options -> General -> Graphics.

- In Quarto add `ragg_png` as the chunk output device:
:::

:::{.fragment}
```{r}
#| echo: true
#| eval: false
---
knitr:
  opts_chunk:
    dev: "ragg_png"
---
```
:::

## Getting fonts

To ensure your work is reproducible, you probably want to make sure you use an open font that is available to everyone.

One of the easiest ways to do this is to use a font repository online such as Google fonts or FontSquirrel.

The functions `get_from_google_fonts()`, `get_from_font_squirrel()` and `require_font()` are available for this.

`require_font()` checks if you already have the font, and if not, will download it.

```{r}
#| echo: true
library(systemfonts)
require_font("Ubuntu")
```

## Using the font

```{r}
#| echo: true
#| fig-width: 10
#| fig.height: 5
#| fig.align: 'center'
require_font("Ubuntu")
g1 +
  theme_minimal(base_family="Ubuntu", base_size=20)
```

## Using a font

```{r}
#| echo: true
#| fig-width: 10
#| fig.height: 5
#| fig.align: 'center'
require_font("Barrio")
g1 +
  theme_minimal(base_family="Barrio", base_size=20)
```

## Using different font weights

Some fonts have different weights available: more than just plain and bold:

```{r}
#| echo: true
system_fonts() |> filter(family == "Ubuntu") |>
  select(name, style, weight, width, path)
```

## Using different font weights

We can register these as variants so we can use them within the restricted plain vs bold system in R:

```{r}
#| echo: true
register_variant(
  "Ubuntu Light",
  "Ubuntu",
  weight='light'
)
font_info("Ubuntu Light") |> select(family, style, name, weight)
```

## Using different font weights

:::: .columns

:::{.column}
```{r}
#| echo: true
#| fig-width: 6
#| fig.height: 4
#| fig.align: 'center'
g1 +
  theme_minimal(base_family="Ubuntu", base_size=20)
```

:::

:::{.column}
```{r}
#| echo: true
#| fig-width: 6
#| fig.height: 4
#| fig.align: 'center'
g1 +
  theme_minimal(base_family="Ubuntu Light", base_size=20)
```
:::

::::

# Figure design

## Original

```{r}
vacc <- read_csv(here::here("data/covid_vacc.csv")) |>
  filter(Dose == "Fully vaccinated")
theme_set(theme_grey(base_size=20))
```

::::{.columns}

:::{.column}
```{r design1}
#| echo: true
#| eval: false
ggplot(vacc) +
  aes(x=Prop, y=DHB, col=Age) +
  geom_point()
```

:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label="design1"
```
:::

::::

## Reverse Y

::::{.columns}

:::{.column}
```{r design2}
#| echo: true
#| eval: false
ggplot(vacc) +
  aes(x=Prop, y=fct_rev(DHB), col=Age) +
  geom_point()
```

:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label="design2"
```
:::

::::

## Clean labels

::::{.columns}

:::{.column}
```{r design3}
#| echo: true
#| eval: false
ggplot(vacc) +
  aes(x=Prop, y=fct_rev(DHB), col=Age) +
  geom_point() +
  labs(x = NULL, y = NULL)
```

:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label="design3"
```
:::

::::

## Percent scale{.smaller}

::::{.columns}

:::{.column}
```{r design4}
#| echo: true
#| eval: false
ggplot(vacc) +
  aes(x=Prop, y=fct_rev(DHB), col=Age) +
  geom_point() +
  labs(x = NULL, y = NULL) +
  scale_x_continuous(labels = scales::label_percent())
```

:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label="design4"
```
:::

::::

## Theme{.smaller}

::::{.columns}

:::{.column}
```{r design5}
#| echo: true
#| eval: false
ggplot(vacc) +
  aes(x=Prop, y=fct_rev(DHB), col=Age) +
  geom_point() +
  labs(x = NULL, y = NULL) +
  scale_x_continuous(labels = scales::label_percent()) +
  theme_minimal(base_size=16)
```

:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label="design5"
```
:::

::::

## Guide position{.smaller}

::::{.columns}

:::{.column}
```{r design6}
#| echo: true
#| eval: false
ggplot(vacc) +
  aes(x=Prop, y=fct_rev(DHB), col=Age) +
  geom_point() +
  labs(x = NULL, y = NULL) +
  scale_x_continuous(labels = scales::label_percent()) +
  theme_minimal(base_size=16) +
  theme(legend.position='bottom',
        legend.margin=margin()) # zero margin
```

:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label="design6"
```
:::

::::

## Point shape/colour{.smaller}

::::{.columns}

:::{.column}
```{r design7}
#| echo: true
#| eval: false
ggplot(vacc) +
  aes(x=Prop, y=fct_rev(DHB), shape=Age, fill=Prop >= 0.9) +
  geom_point() +
  labs(x = NULL, y = NULL, shape = NULL) +
  scale_shape_manual(values=21:24) +
  scale_x_continuous(labels = scales::label_percent()) +
  scale_fill_manual(values = c('pink', 'steelblue')) +
  guides(fill = 'none') +
  theme_minimal(base_size=16) +
  theme(legend.position='bottom',
        legend.margin=margin()) # zero margin
```

:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label="design7"
```
:::

::::

## 90% goal{.smaller}

::::{.columns}

:::{.column}
```{r design8}
#| echo: true
#| eval: false
ggplot(vacc) +
  aes(x=Prop, y=fct_rev(DHB), shape=Age, fill=Prop >= 0.9) +
  geom_point() +
  geom_vline(xintercept=0.9, col='grey30')+
  labs(x = NULL, y = NULL, shape = NULL) +
  scale_shape_manual(values=21:24) +
  scale_x_continuous(labels = scales::label_percent()) +
  scale_fill_manual(values = c('pink', 'steelblue')) +
  guides(fill = 'none') +
  theme_minimal(base_size=16) +
  theme(legend.position='bottom',
        legend.margin=margin()) # zero margin
```

:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label="design8"
```
:::

::::

## Axis range and space{.smaller}

::::{.columns}

:::{.column}
```{r design9}
#| echo: true
#| eval: false
ggplot(vacc) +
  aes(x=Prop, y=fct_rev(DHB), shape=Age, fill=Prop >= 0.9) +
  geom_vline(xintercept=0.9, col='grey30')+
  geom_point() +
  labs(x = NULL, y = NULL, shape = NULL) +
  scale_shape_manual(values=21:24) +
  scale_x_continuous(labels = scales::label_percent(),
                     limits=c(0,1),
                     expand=c(0,0)) +
  scale_fill_manual(values = c('pink', 'steelblue')) +
  guides(fill = 'none') +
  theme_minimal(base_size=20) +
  theme(legend.position='bottom',
        legend.margin=margin()) # zero margin
```

:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label="design9"
```
:::

::::

## Link points{.smaller}

::::{.columns}

:::{.column}
```{r design10}
#| echo: true
#| eval: false
ggplot(vacc) +
  aes(x=Prop, y=fct_rev(DHB)) +
  geom_line(col="grey70") +
  geom_vline(xintercept=0.9, col='grey30')+
  geom_point(mapping=aes(shape=Age, fill=Prop >= 0.9)) +
  labs(x = NULL, y = NULL, shape = NULL) +
  scale_shape_manual(values=21:24) +
  scale_x_continuous(labels = scales::label_percent(),
                     limits=c(0,1),
                     expand=c(0,0)) +
  scale_fill_manual(values = c('pink', 'steelblue')) +
  guides(fill = 'none') +
  theme_minimal(base_size=16) +
  theme(legend.position='bottom',
        legend.margin=margin()) # zero margin
```

:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label="design10"
```
:::

::::


## Remove panel grid{.smaller}

::::{.columns}

:::{.column}
```{r design11}
#| echo: true
#| eval: false
ggplot(vacc) +
  aes(x=Prop, y=fct_rev(DHB)) +
  geom_line(col="grey70") +
  geom_vline(xintercept=0.9, col='grey30')+
  geom_point(mapping=aes(shape=Age, fill=Prop >= 0.9)) +
  labs(x = NULL, y = NULL, shape = NULL) +
  scale_shape_manual(values=21:24) +
  scale_x_continuous(labels = scales::label_percent(),
                     limits=c(0,1),
                     expand=c(0,0)) +
  scale_fill_manual(values = c('pink', 'steelblue')) +
  guides(fill = 'none') +
  theme_minimal(base_size=16) +
  theme(legend.position='bottom',
        legend.margin=margin(), # zero margin
        panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_line(colour='grey90', size=0.5))
```

:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label="design11"
```
:::

::::


## Axis size/colour{.smaller}

::::{.columns}

:::{.column}
```{r design12}
#| echo: true
#| eval: false
ggplot(vacc) +
  aes(x=Prop, y=fct_rev(DHB)) +
  geom_line(col="grey70") +
  geom_vline(xintercept=0.9, col='grey30')+
  geom_point(mapping=aes(shape=Age, fill=Prop >= 0.9)) +
  labs(x = NULL, y = NULL, shape = NULL) +
  scale_shape_manual(values=21:24) +
  scale_x_continuous(labels = scales::label_percent(),
                     limits=c(0,1),
                     expand=c(0,0)) +
  scale_fill_manual(values = c('pink', 'steelblue')) +
  guides(fill = 'none') +
  theme_minimal(base_size=16) +
  theme(legend.position='bottom',
        legend.margin=margin(), # zero margin
        panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_line(color='grey90', size=0.5),
        axis.text.x = element_text(size=rel(0.8), colour='grey70'))
```

:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label="design12"
```
:::

::::

## Add 90% label{.smaller}

::::{.columns}

:::{.column}
```{r design13}
#| echo: true
#| eval: false
ggplot(vacc) +
  aes(x=Prop, y=fct_rev(DHB)) +
  geom_line(col="grey70") +
  geom_vline(xintercept=0.9, col='grey30')+
  geom_point(mapping=aes(shape=Age, fill=Prop >= 0.9)) +
  annotate(geom='text', x=0.9, y=21, label="90%", size=7) +
  labs(x = NULL, y = NULL, shape = NULL) +
  scale_shape_manual(values=21:24) +
  scale_x_continuous(labels = scales::label_percent(),
                     limits=c(0,1),
                     expand=c(0,0)) +
  scale_fill_manual(values = c('pink', 'steelblue')) +
  coord_cartesian(clip='off') +
  guides(fill = 'none') +
  theme_minimal(base_size=16) +
  theme(legend.position='bottom',
        legend.margin=margin(), # zero margin
        panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_line(colour='grey90', size=0.5),
        axis.text.x = element_text(size=rel(0.8), colour='grey70'))
```

:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label="design13"
```
:::

::::


## Add title{.smaller}

::::{.columns}

:::{.column}
```{r design14}
#| echo: true
#| eval: false
ggplot(vacc) +
  aes(x=Prop, y=fct_rev(DHB)) +
  geom_line(col="grey70") +
  geom_vline(xintercept=0.9, col='grey30')+
  geom_point(mapping=aes(shape=Age, fill=Prop >= 0.9)) +
  annotate(geom='text', x=0.9, y=21, label="90%", size=7) +
  labs(x = NULL, y = NULL, shape = NULL,
       title = "The 90% project: Covid-19 vaccination rates by DHB") +
  scale_shape_manual(values=21:24) +
  scale_x_continuous(labels = scales::label_percent(),
                     limits=c(0,1),
                     expand=c(0,0)) +
  scale_fill_manual(values = c('pink', 'steelblue')) +
  coord_cartesian(clip='off') +
  guides(fill = 'none') +
  theme_minimal(base_size=16) +
  theme(legend.position='bottom',
        legend.margin=margin(), # zero margin
        panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_line(colour='grey90', size=0.5),
        axis.text.x = element_text(size=rel(0.8), colour='grey70'),
        plot.title.position = 'plot')
```

:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label="design14"
```
:::

::::

## Plot margins{.smaller}

::::{.columns}

:::{.column}
```{r design15}
#| echo: true
#| eval: false
ggplot(vacc) +
  aes(x=Prop, y=fct_rev(DHB)) +
  geom_line(col="grey70") +
  geom_vline(xintercept=0.9, col='grey30')+
  geom_point(mapping=aes(shape=Age, fill=Prop >= 0.9)) +
  annotate(geom='text', x=0.9, y=21, label="90%", size=7) +
  labs(x = NULL, y = NULL, shape = NULL,
       title = "The 90% project: Covid-19 vaccination rates by DHB") +
  scale_shape_manual(values=21:24) +
  scale_x_continuous(labels = scales::label_percent(),
                     limits=c(0,1),
                     expand=c(0,0)) +
  scale_fill_manual(values = c('pink', 'steelblue')) +
  coord_cartesian(clip='off') +
  guides(fill = 'none') +
  theme_minimal(base_size=16) +
  theme(legend.position='bottom',
        legend.margin=margin(), # zero margin
        panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_line(colour='grey90', size=0.5),
        axis.text.x = element_text(size=rel(0.8), colour='grey70'),
        plot.title.position = 'plot',
        plot.margin = margin(6, 12, 6, 6)) # more on right
```

:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label="design15"
```
:::

::::

## Add caption{.smaller}

::::{.columns}

:::{.column}
```{r design16}
#| echo: true
#| eval: false
ggplot(vacc) +
  aes(x=Prop, y=fct_rev(DHB)) +
  geom_line(col="grey70") +
  geom_vline(xintercept=0.9, col='grey30')+
  geom_point(mapping=aes(shape=Age, fill=Prop >= 0.9)) +
  annotate(geom='text', x=0.9, y=21, label="90%", size=7) +
  labs(x = NULL, y = NULL, shape = NULL,
       title = "The 90% project: Covid-19 vaccination rates by DHB",
       caption = "Data from Ministry of Health. Chart by Jonathan Marshall") +
  scale_shape_manual(values=21:24) +
  scale_x_continuous(labels = scales::label_percent(),
                     limits=c(0,1),
                     expand=c(0,0)) +
  scale_fill_manual(values = c('pink', 'steelblue')) +
  coord_cartesian(clip='off') +
  guides(fill = 'none') +
  theme_minimal(base_size=16) +
  theme(legend.position='bottom',
        legend.margin=margin(), # zero margin
        panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_line(colour='grey90', size=0.5),
        axis.text.x = element_text(size=rel(0.8), colour='grey70'),
        plot.title.position = 'plot',
        plot.margin = margin(6, 12, 6, 6), # more space top and right
        plot.caption = element_text(size=rel(0.6), colour='grey50'))
```

:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label="design16"
```
:::

::::


## Change font{.smaller}

::::{.columns}

:::{.column}
```{r design17}
#| echo: true
#| eval: false
require_font('Source Sans 3')
ggplot(vacc) +
  aes(x=Prop, y=fct_rev(DHB)) +
  geom_line(col="grey70") +
  geom_vline(xintercept=0.9, col='grey30')+
  geom_point(mapping=aes(shape=Age, fill=Prop >= 0.9)) +
  annotate(geom='text', x=0.9, y=21, label="90%", size=7) +
  labs(x = NULL, y = NULL, shape = NULL,
       title = "The 90% project: Covid-19 vaccination rates by DHB",
       caption = "Data from Ministry of Health. Chart by Jonathan Marshall") +
  scale_shape_manual(values=21:24) +
  scale_x_continuous(labels = scales::label_percent(),
                     limits=c(0,1),
                     expand=c(0,0)) +
  scale_fill_manual(values = c('pink', 'steelblue')) +
  coord_cartesian(clip='off') +
  guides(fill = 'none') +
  theme_minimal(base_family="Source Sans 3", base_size=18) +
  theme(legend.position='bottom',
        legend.margin=margin(), # zero margin
        panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_line(colour='grey90', size=0.5),
        axis.text.x = element_text(size=rel(0.8), colour='grey70'),
        plot.title.position = 'plot',
        plot.title = element_text(face='bold'),
        plot.margin = margin(6, 12, 6, 6), # more space top and right
        plot.caption = element_text(size=rel(0.6), colour='grey50'))
```

:::

:::{.column .absolute top=0 right=-5%}
```{r}
#| ref.label="design17"
```
:::

::::

## Further reading

- [Themes](https://ggplot2-book.org/themes.html) ggplot2: Elegant Graphics for Data Analysis

- [Changing the Appearance of Plots with Themes](http://vissoc.co/08-polishing.html#change-the-appearance-of-plots-with-themes) from Data Visualisation - A Practical Introduction

- [Multi-panel figures](https://clauswilke.com/dataviz/multi-panel-figures.html) from Fundamentals of Data Visualisation

- [Workhorses, Show Ponies, Uniconrs](http://vissoc.co/08-polishing.html#workhorses-show-ponies-unicorns) from Data Visualisation - A Practical Introduction

- [Balance the data and the context](https://clauswilke.com/dataviz/balance-data-context.html) from Fundamentals of Data Visualisation

- The [patchwork](https://patchwork.data-imaginist.com) package.
