---
title: "Points in Space!"
format:
  revealjs:
    theme: [solarized, slides.scss]
    width: 1280
    height: 720
    margin: 0.1
    echo: true
    message: false
    fig-width: 7
    fig-height: 7
---

```{r}
#| echo: false

library(tidyverse)
theme_set(theme_minimal(base_size=20))
```

# Spatial data

## Spatial data

Spatial data is anything that has some position or coordinates associated with it.

There are four main types.

::: {.fragment}
**Points**: x, y coordinates possibly with some other data associated with them.
:::

::: {.fragment}
**Lines**: A linked set of points that may have data on the links.
:::

::: {.fragment}
**Polygons**: closed polygons that describe the boundaries of a set of regions which possibly have some other data associated with them.
:::

::: {.fragment}
**Grids**: A special case where the polygons are uniform rectangles that form a grid.
:::

::: {.fragment}
::: aside
We'll be looking at points, polygons and grids. Lines will be present through polygon borders though.
:::
:::

## Points

Point data are the locations of observations through some spatial region.

The spatial region might be bounded (e.g. points within a country or city) or unbounded.

Can be more than two dimensional (e.g. points in 3D space) but we'll be focusing on 2D, and will generally assume that this 2D data are **planar**.

:::{.fragment}
Many such data that we assume are planar are not! e.g. locations on the planet are better approximated by a sphere but like the term "plane sailing" we'll temporally become flat-earthers.
:::

## Questions about points

The simplest question about points is: "why are they there?"

i.e. what is the **process** that gave rise to this particular pattern of points that we see.

:::{.fragment}
Other questions may be:

- Is something we've measured at these points related to the location? *Spatial correlation*

- Can we take a measure from one set of points and infer what that measure might be somewhere else? *Spatial interpolation*
:::

## Earthquakes around Fiji

:::: {.columns}

::: {.column}
```{r quakes1}
#| eval: false
ggplot(quakes) +
  aes(x=long, y=lat) +
  geom_point(alpha=0.2)
```

- Have longitude/latitude information

- Assume the data are 'flat' in these

:::

::: {.column .absolute top=0 right=-5%}
```{r}
#| echo: false
#| ref.label: "quakes1"
```
:::

::::

## Earthquakes around Fiji

:::: {.columns}

::: {.column}
```{r quakes2}
#| eval: false
ggplot(quakes) +
  aes(x=long, y=lat) +
  geom_point(alpha=0.2) +
  coord_fixed()
```

- Have longitude/latitude information

- Assume the data are 'flat' in these

- The aspect ratio is important though!

:::

::: {.column .absolute top=0 right=-5%}
```{r}
#| echo: false
#| ref.label: "quakes2"
```
:::

::::

## Cricket wagon-wheels

::::{.columns}

:::{.column}

Brendon McCullum's innings of
302 against India in 2014

:::

::: {.column .absolute top=0 right=-5%}

```{r}
#| echo: false
nz_cric <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/cricket/nz_bat_scores.csv")

bmc <- nz_cric |> filter(p_match == 667653,
                  score != 0, inns == 3,
                  wagonX != 0, wagonY != 0,
                  bat == "Brendon McCullum")

oval <- tibble(theta = seq(0, 2*pi, length.out=200)) |>
  mutate(x = 175*sin(theta), y = 175*cos(theta))

ggplot(bmc) +
  geom_polygon(data=oval, aes(x=x, y=y), fill='green4', col=NA) +
  geom_segment(aes(x=wagonX, y=wagonY, colour=outcome, linewidth=outcome, xend=0, yend=0)) +
  scale_colour_manual(values = c(six = 'red2', four='gold', run='white')) +
  scale_linewidth_manual(values= c(six = 1.5, four=1, run=0.5)) +
  theme_void() + 
#  theme(panel.background=element_rect(fill='green4')) +
  guides(colour='none', linewidth='none') +
  coord_fixed()
```

:::
::::

## Cricket wagon-wheels

::::{.columns}

:::{.column}

```{r wagon1}
#| eval: false
ggplot(bmc) +
  aes(x=wagonX, y=wagonY) +
  geom_point(alpha=0.3) +
  coord_fixed()
```

Brendon McCullum's innings of
302 against India in 2014 as a point pattern.

:::
::: {.column .absolute top=0 right=-5%}
```{r}
#| echo: false
#| ref.label: "wagon1"
```
:::

::::

# Point processes

## Point processes

A point process is the statistical model that gives rise to the location of points.

:::{.incremental}
- A one-dimensional point process produces event times.

- A two-dimensional point process produces event locations.
:::

:::{.fragment}
The simplest type of model is a *Poisson process* where we assume events occur independently at a given rate.
:::

## Poisson processes

A spatial *Poisson process* generates $N$ points in an area $A$ according to the area of the region and rate $\lambda$.

:::{.fragment}
For any region $A$ and rate $\lambda$ we expect the number of points $N(A)$ to follow a Poisson distribution:

$$
P(N(A) = n) \sim \textsf{Poisson}(\lambda|A|)
$$
where $|A|$ is the area of region $A$.
:::

:::{.fragment}
The **strong** assumption is that the points don't interact: i.e. their location is not influenced by the location of other points.
:::

## What is random?

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 6
#| fig-align: 'center'
library(spatstat)

set.seed(2025)
csr <- rpoispp(lambda=50, nsim=1) |> as.data.frame()
tom <- rThomas(8, 0.08, 7) |> as.data.frame()
mat <- rMaternII(100, r=0.07) |> as.data.frame()
sims <- bind_rows(csr, tom, mat, .id='model')
ggplot(sims) +
  aes(x=x, y=y) +
  geom_point() +
  theme_void() +
  theme(panel.border = element_rect(colour='black', linewidth=0.5),
        strip.text = element_blank()) +
  facet_wrap(vars(model))
```

## Homogeneous processes

- If the **intensity** $\lambda$ is constant across the region, then the process is **homogeneous**.

- The chance of a point being in any location is fully randomised.

- The point patterns that result exhibit **complete spatial randomness** or CSR.

:::{.fragment}
This results in a simple statistical model, and works well-enough in a surprising number of examples.
:::

:::{.fragment}
As a result there are various tests to check whether a point pattern is CSR.
:::
## Inhomogeneous processes

A more realistic process is the **inhomogeneous** Poisson process where the rate $\lambda = \lambda(\mathbf{x})$ depends on the location $\mathbf{x}$.

The number of points $N(A)$ in region $A$ still follows a Poisson distribution, but the intensity $\Lambda$ now depends on $\mathbf{x}$.

$$
P(N(A) = n) \sim \textsf{Poisson}(\Lambda(A)), \qquad \Lambda(A) = \int_A \lambda(x) dx
$$

## Points are still independent

:::{.incremental}
- In an inhomogeneous Poisson process the points still don't interact with each other.

- Having more people nearby does not mean that additional people are likely to be there.

- It's just that there's some external pressure (e.g. "the sea") from the intensity $\lambda$ that results in more points in one place than another.

- Other fancier processes that allow for **spatial clustering** or **inhibition** are available.

- Often the inhomogeneous Poisson process does pretty well though!
:::

## What is important in a point process?

- The key thing of importance is the underlying **intensity** function $\lambda(\mathbf{x})$.

- $\lambda(\mathbf{x})$ contains two separate ideas: How many points are generated (magnitude) plus where they are (density).

- It is the **density** that is of interest: where are points more or less likely to be?

- To estimate the density we use a kernel density estimator.

- Same idea as `geom_density()` extended to two dimensions.

## Estimating point intensity

We let $\lambda(\mathbf{x}) = \lambda_0 f(\mathbf{x})$ where $\lambda_0$ represents magnitude (number of points to be generated), and $f(\mathbf{x})$ is a probability density function (where they go).

:::{.fragment}
We estimate $f(\mathbf{x})$ using a KDE in two dimensions:

$$
\hat f(\mathbf{x})= \frac{1}{n} \sum_{i=1}^n K_h(\mathbf{x} - \mathbf{x}_i)
$$

where $\mathbf{x}$ and $\mathbf{x}_i$ are 2-d spatial locations, and the kernel $K_h$ is a two dimensional kernel.
:::

:::{.fragment}
We usually set $K_h$ to be a bivariate normal distribution and assume that it is rotationally symmetric (isotropic) so that $h$ is a scalar.
:::

:::{.fragment}
In `ggplot2` we can use `geom_density_2d().`
:::


## Earthquakes around Fiji

:::: {.columns}

::: {.column}
```{r quakes3}
#| eval: false
ggplot(quakes) +
  aes(x=long, y=lat) +
  geom_point(alpha=0.2) +
  coord_fixed()
```

:::

::: {.column .absolute top=0 right=-5%}
```{r}
#| echo: false
#| ref.label: "quakes3"
```
:::

::::

## Earthquakes around Fiji

:::: {.columns}

::: {.column}
```{r quakes4}
#| eval: false
ggplot(quakes) +
  aes(x=long, y=lat) +
  geom_point(alpha=0.2) +
  geom_density_2d() +
  coord_fixed()
```

Use `geom_density_2d()` for a
density estimator
:::

::: {.column .absolute top=0 right=-5%}
```{r}
#| echo: false
#| ref.label: "quakes4"
```
:::

::::


## Earthquakes around Fiji

:::: {.columns}

::: {.column}
```{r quakes5}
#| eval: false
ggplot(quakes) +
  aes(x=long, y=lat) +
  geom_point(alpha=0.2) +
  geom_density_2d(adjust=0.5) +
  coord_fixed()
```

Use `geom_density_2d()` for a
density estimator.

Can tune the bandwidth with `adjust`
:::

::: {.column .absolute top=0 right=-5%}
```{r}
#| echo: false
#| ref.label: "quakes5"
```
:::

::::


## Earthquakes around Fiji

:::: {.columns}

::: {.column}
```{r quakes6}
#| eval: false
ggplot(quakes) +
  aes(x=long, y=lat) +
  geom_density_2d_filled(adjust=0.5) +
  geom_point(alpha=0.2, size=0.2) +
  scale_fill_viridis_d() +
  guides(fill='none') +
  coord_fixed()
```

Use `geom_density_2d_filled()` to
fill the contours.
:::

::: {.column .absolute top=0 right=-5%}
```{r}
#| echo: false
#| ref.label: "quakes6"
```
:::

::::

## Cricket

::::{.columns}

:::{.column}

```{r wagon2}
#| eval: false
ggplot(bmc) +
  aes(x=wagonX, y=wagonY) +
  geom_point(alpha=0.3) +
  coord_fixed()
```

Brendon McCullum's innings of
302 against India in 2014 as a point pattern.

:::
::: {.column .absolute top=0 right=-5%}
```{r}
#| echo: false
#| ref.label: "wagon2"
```
:::

::::

## Cricket

::::{.columns}

:::{.column}

```{r wagon3}
#| eval: false
ggplot(bmc) +
  aes(x=wagonX, y=wagonY) +
  geom_point(alpha=0.3) +
  geom_density_2d() +
  coord_fixed()
```

Brendon McCullum's innings of
302 against India in 2014.

:::
::: {.column .absolute top=0 right=-5%}
```{r}
#| echo: false
#| ref.label: "wagon3"
```
:::

::::

## Cricket

```{r}
#| echo: false
sangakkara <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/cricket/sangakkara.csv")
```
::::{.columns}

:::{.column}

```{r wagon5}
#| eval: false
ggplot(sangakkara) +
  aes(x=wagonX, y=wagonY) +
  geom_point(alpha=0.3) +
  geom_density_2d() +
  coord_fixed()
```

Kumar Sangakkara's 319 against
Bangladesh in 2014.

:::
::: {.column .absolute top=0 right=-5%}
```{r}
#| echo: false
#| ref.label: "wagon5"
```
:::

::::

## KDEs give uniform smoothing

:::{.incremental}
- Our kernel density estimator uses a single bandwidth $h$ for the smoothing.

- This applies across the entire region.

- But, if the intensity is changing across the area, then that implies that the amount of data (information!) we have differs across the region.

- If we smooth everything by the same amount, places where data are dense will be tend to be *over-smoothed*, whilst places where data are sparse may be *under-smoothed*.

- Ideally we'd allow our estimator to adapt the amount of smoothing to the situation.
:::

## Adaptive KDE

- In adaptive kernel density estimates we allow $h = h(\mathbf{x})$ to vary with $\mathbf{x}$.

- Where data are sparse ($f(\mathbf{x})$ is small) we want more smoothing.

- Where data are dense ($f(\mathbf{x})$ is large) we want less smoothing.

- Makes sense to smooth in proportion to $\frac{1}{\sqrt{f(\mathbf{x})}}$. **Why this form?**

:::{.fragment}
Chicken and egg problem: We need $h$ to estimate $f$ and $f$ to estimate $h$.
:::

## Iterating out of the chicken/egg

:::{.incremental}
- First get a **pilot estimate** $\hat f_p(x)$ with using constant smoothing with a fixed $h$.

- Set $h(x) = \frac{1}{\sqrt{f_p(x)}}$ and use that in a second round of estimation[^2].

- We could iterate again, using our adaptive estimate to recompute $h(x)$ and estimate a third time, but it's usually not necessary.
:::

[^2]: See [Abramson 1982](https://projecteuclid.org/journals/annals-of-statistics/volume-10/issue-4/On-Bandwidth-Variation-in-Kernel-Estimates-A-Square-Root-Law/10.1214/aos/1176345986.full) if you're especially keen on the maths!

## Kumar Sangakkara

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 6
#| fig-align: 'center'
sangakkara.ppp <- as.ppp(sangakkara |> select(wagonX, wagonY),
                         W = owin(c(-200,200), c(-200,200)))
h0=bw.ppl(sangakkara.ppp)

fixed <-density(sangakkara.ppp, sigma=h0) |> as.data.frame()
adaptive <- densityAdaptiveKernel(sangakkara.ppp, h0=h0) |> as.data.frame()

bind_rows(lst(fixed, adaptive), .id="model") |>
  mutate(model = fct_relevel(model, 'fixed')) |>
  ggplot() +
  geom_point(data=sangakkara, mapping=aes(x=wagonX, y=wagonY), alpha=0.3) +
  geom_contour(mapping=aes(x=x, y=y, z=value)) +
  coord_fixed() +
  facet_wrap(vars(model)) +
  theme(panel.spacing.x = unit(20, units='mm'))
```

## Brendon McCullum

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 6
#| fig-align: 'center'
bmc.ppp <- as.ppp(bmc |> select(wagonX, wagonY),
                         W = owin(c(-200,200), c(-200,200)))
h0=bw.ppl(bmc.ppp)

fixed <-density(bmc.ppp, sigma=h0) |> as.data.frame()
adaptive <- densityAdaptiveKernel(bmc.ppp, h0=h0) |> as.data.frame()

bind_rows(lst(fixed, adaptive), .id="model") |>
  mutate(model = fct_relevel(model, 'fixed')) |>
  ggplot() +
  geom_point(data=bmc, mapping=aes(x=wagonX, y=wagonY), alpha=0.3) +
  geom_contour(mapping=aes(x=x, y=y, z=value)) +
  coord_fixed() +
  facet_wrap(vars(model)) +
  theme(panel.spacing.x = unit(20, units='mm'))
```

## Adaptive quakes

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 6
#| fig-align: 'center'
quakes.ppp <- as.ppp(quakes |> select(long, lat),
                         W = owin(range(quakes$long), range(quakes$lat)))
h0=bw.ppl(quakes.ppp)

fixed <-density(quakes.ppp, sigma=h0) |> as.data.frame()
adaptive <- densityAdaptiveKernel(quakes.ppp, h0=h0) |> as.data.frame()

bind_rows(lst(fixed, adaptive), .id="model") |>
  mutate(model = fct_relevel(model, 'fixed')) |>
  mutate(value = if_else(value < 1e-10, 0, value)) |>
  ggplot() +
  geom_point(data=quakes, mapping=aes(x=long, y=lat), alpha=0.1) +
  geom_contour(mapping=aes(x=x, y=y, z=value)) +
  coord_fixed() +
  facet_wrap(vars(model)) +
  theme(panel.spacing.x = unit(20, units='mm'))
```
::: aside
More of this in 233.214 GIS and Spatial Statistics
:::

## Adding context to quakes

::::{.columns}

:::{.column}

```{r nzquake}
#| eval: false
nz <- map_data('nz')
ggplot(quakes) +
  aes(x=long, y=lat) +
  geom_point(alpha=0.3) +
  geom_path(data=nz,
            aes(group=group)) +
  coord_fixed()
```

Can plot NZ on the same map

:::
::: {.column .absolute top=0 right=-5%}
```{r}
#| echo: false
#| ref.label: "nzquake"
```
:::

::::

## Adding context to quakes

::::{.columns}

:::{.column}

```{r nzquake2}
#| eval: false
nz <- map_data('nz')
ggplot(quakes) +
  aes(x=long, y=lat) +
  geom_point(alpha=0.3) +
  geom_path(data=nz,
            aes(group=group)) +
  coord_fixed()
```

But it's not plane sailing anymore!

:::
::: {.column .absolute top=0 right=-5%}
```{r}
#| echo: false
library(sf)
dists <- quakes |>
  reframe(long=range(long), lat=range(lat, nz$lat)) |>
  complete(long, lat) |>
  st_as_sf(coords = c('long', 'lat'), remove=FALSE) |>
  st_set_crs(4326)

d <- st_distance(dists)

# hacky hacky
geodist <- tribble(~line, ~x1, ~x2, ~y1, ~d,
        1, 166, 188, -47.4, d[1,3],
        2, 166, 188, -10.7, d[4,2])

ggplot(quakes) +
  aes(x=long, y=lat) +
  geom_point(alpha=0.3) +
  geom_path(data=nz, aes(group=group), col='grey40') +
  geom_segment(data=geodist, aes(x=x1, xend=x2, y=y1), col='darkred',
               arrow = arrow(angle=10, type='closed', ends='both')) +
  geom_text(data=geodist, aes(x=(x1+x2)/2, y=y1, label=paste(round(d/1000), 'km')),
            col='darkred', vjust=-.2) +
  coord_fixed()
```
:::

::::

## Further reading

- [Maps](https://ggplot2-book.org/maps.html) from ggplot2: Elegant Graphics for Data Analysis

- [Types of spatial data](https://www.paulamoraga.com/book-spatial/types-of-spatial-data.html) in Spatial Statistics for Data Science: Theory and Practice with R

- Take 233.214 GIS and Spatial Statistics!

