---
title: "Portfolio 1: Plastic, plastic, everywhere"
format:
  html:
    embed-resources: true
---

## Introduction

Recall from Workshop 1 that LEGO enthusiasts have noted that sets have increased detail and complexity which results in more parts but potentially similar amounts of plastic.

In this portfolio your goal is to produce a figure to demonstrate whether the amount of plastic in LEGO sets for a given price is changing over time.

We use data on `lego_sets` and `lego_parts` joined together:

```{r}
#| message: false
#| echo: false

library(tidyverse)

lego_sets <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/lego/lego_sets.csv")
lego_parts <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/lego/lego_parts.csv.gz")

set_weights <- lego_parts |>
  group_by(set) |>
  summarise(weight = sum(weight*quantity, na.rm=TRUE))

sets_with_weight <- lego_sets |>
  left_join(set_weights, by=join_by(set))
```

## Requirements

You must use the `sets_with_weight` data to produce a chart of price per gram for each lego set for each theme for each year. You should model it on the chart we did in workshop 1 which related `price_per_part` to `year` and `theme`.

 - Ensure you have a facetted plot (facet by theme is recommended).
 - Show both the data (all sets) and a trend on each facet.
 - Your chart should have appropriate labels, a title and subtitle.
 - Use the title to help tell the story of the chart.
 - Your chart should utilise a different colour in each facet. You don't need a guide for this.
 - The colours you choose should be different to the defaults. You may utilise a palette or choose your own.
 - You should submit this .qmd file. You should ensure that it renders correctly before submission.
 - You will be graded both on the appearance of the figure and your code.

## Your Figure

```{r}
# replace this comment with your figure code
```
